---
title: "Untitled"
output: html_document
date: "2026-01-06"
---

# Covariate Preparation Pipeline for 5 km Under‑Five Mortality Mapping in (2000‑2015)

IMPORTANT – RAW DATA NOT INCLUDED, BUT PROCESSED OUTPUTS FOR ALL YEARS (2000‑2000) ARE PROVIDED

This script processes freely available raw rasters (ERA5, GBD, MODIS, WorldPop, Malaria Atlas Project, etc.). These raw data are NOT included in this deposit due to their large size. The folder paths (e.g., /...) reflect the author’s local machine; users must download the raw data themselves and adjust the paths if they wish to re‑run the script.

However, the final outputs of this script – adjusted covariates (21 variables reduced to 16 after correlation filtering), 5 km population rasters, and inhabited area masks for ALL years 2000‑2000 – ARE included in this deposit under the `adjusted_covariates/` folder. These ready‑to‑use GeoTIFFs can be directly employed in downstream analyses (e.g., censoring adjustment or INLA modelling) without re‑running the script.

The script is shared for methodological transparency and to document every processing step (log1p, z‑score, capping, metadata generation). Users who wish to reproduce the entire workflow or process additional years can do so after obtaining the raw data and modifying the file paths.

```{r}
library(terra)
library(e1071)
library(jsonlite)
library(dplyr)
library(caret)    
library(corrplot)  
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
eth_shp <- vect("D:/gadm41_ETH_0.shp") 
target_crs <- "EPSG:32637"
eth_utm <- project(eth_shp, target_crs)
e <- ext(eth_utm)
template_5km <- rast(e, resolution = 5000, crs = target_crs)
values(template_5km) <- 1:ncell(template_5km)
template_5km <- mask(template_5km, eth_utm)
template <- mask(template_5km, eth_utm)
```

# Covariate Preparation Pipeline for 5 km Under‑Five Mortality Mapping in Ethiopia (2000‑2000)

# Dignosis for transformation decation 

```{r}
folder_path <- "D:/dewpoi_temp_2m" # call 
output_path <- "D:/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (dewPoi_2000.tif to dewPoi_1tif)
files <- sprintf("dewPoi_%02d.tif", 0:19)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_dewpoint <- rast(rasters)
# Pool all values into a single vector
values_dewpoint <- values(stack_dewpoint, mat = FALSE)
values_dewpoint <- values_dewpoint[!is.na(values_dewpoint)]
# Compute diagnostics
mean_val <- mean(values_dewpoint)
median_val <- median(values_dewpoint)
min_val <- min(values_dewpoint)
max_val <- max(values_dewpoint)
skew_val <- skewness(values_dewpoint)
percent_zeros <- sum(values_dewpoint == 0)/length(values_dewpoint) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "dewpoint_temp_2m",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "DewPoint_Diagnosis_2000_201csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
```

```{r}
# Parameters
folder_path <- "D:/deareha"
output_path <- "D:/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (2000–2017 -> dear_et_2000.tif to dear_et_1tif)
files <- sprintf("dear_et_%02d.tif", 0:17)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_deareha <- rast(rasters)
# Pool all values into a single vector
values_deareha <- values(stack_deareha, mat = FALSE)
values_deareha <- values_deareha[!is.na(values_deareha)]
# Compute diagnostics
mean_val <- mean(values_deareha)
median_val <- median(values_deareha)
min_val <- min(values_deareha)
max_val <- max(values_deareha)
skew_val <- skewness(values_deareha)
percent_zeros <- sum(values_deareha == 0)/length(values_deareha) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "deareha",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "Dear_Diagnosis_2000_201csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
```

```{r}
# Parameters
folder_path <- "D:/solarradationdown"
output_path <- "D:/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (rad_2000.tif to rad_1tif)
files <- sprintf("rad_%02d.tif", 0:19)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_solar <- rast(rasters)
# Pool all values into a single vector
values_solar <- values(stack_solar, mat = FALSE)
values_solar <- values_solar[!is.na(values_solar)]
# Compute diagnostics
mean_val <- mean(values_solar)
median_val <- median(values_solar)
min_val <- min(values_solar)
max_val <- max(values_solar)
skew_val <- skewness(values_solar)
percent_zeros <- sum(values_solar == 0)/length(values_solar) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "solarradationdown",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "SolarRad_Diagnosis_2000_201csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
```

```{r}
# Parameters
folder_path <- "D:/potentialevaporation"
output_path <- "D:/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (evap_2000.tif to evap_1tif)
files <- sprintf("evap_%02d.tif", 0:19)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_evap <- rast(rasters)
# Pool all values into a single vector
values_evap <- values(stack_evap, mat = FALSE)
values_evap <- values_evap[!is.na(values_evap)]
# Compute diagnostics
mean_val <- mean(values_evap)
median_val <- median(values_evap)
min_val <- min(values_evap)
max_val <- max(values_evap)
skew_val <- skewness(values_evap)
percent_zeros <- sum(values_evap == 0)/length(values_evap) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "potentialevaporation",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "Evaporation_Diagnosis_2000_201csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
```

```{r}
# Parameters
folder_path <- "D:/runoff"
output_path <- "D:/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (run_2000.tif to run_1tif)
files <- sprintf("run_%02d.tif", 0:19)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_runoff <- rast(rasters)
# Pool all values into a single vector
values_runoff <- values(stack_runoff, mat = FALSE)
values_runoff <- values_runoff[!is.na(values_runoff)]
# Compute diagnostics
mean_val <- mean(values_runoff)
median_val <- median(values_runoff)
min_val <- min(values_runoff)
max_val <- max(values_runoff)
skew_val <- skewness(values_runoff)
percent_zeros <- sum(values_runoff == 0)/length(values_runoff) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "runoff",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "Runoff_Diagnosis_2000_201csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
```

```{r}
# Parameters
folder_path <- "D:/NDVI Ethiopia"
output_path <- "D:/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (NDVI_2000_Ethiopia.tif to NDVI_2019_Ethiopia.tif)
years <- 2000:2019
files <- sprintf("NDVI_%d_Ethiopia.tif", years)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_ndvi <- rast(rasters)
# Pool all values into a single vector
values_ndvi <- values(stack_ndvi, mat = FALSE)
values_ndvi <- values_ndvi[!is.na(values_ndvi)]
# Compute diagnostics
mean_val <- mean(values_ndvi)
median_val <- median(values_ndvi)
min_val <- min(values_ndvi)
max_val <- max(values_ndvi)
skew_val <- skewness(values_ndvi)
percent_zeros <- sum(values_ndvi == 0)/length(values_ndvi) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "NDVI",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "NDVI_Diagnosis_2000_201csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
```

```{r}
# Parameters
folder_path <- "D:/pm5"
output_path <- "D:/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (pm25_et_2000.tif to pm25_et_1tif)
files <- sprintf("pm25_et_%02d.tif", 0:19)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_pm25 <- rast(rasters)
# Pool all values into a single vector
values_pm25 <- values(stack_pm25, mat = FALSE)
values_pm25 <- values_pm25[!is.na(values_pm25)]
# Compute diagnostics
mean_val <- mean(values_pm25)
median_val <- median(values_pm25)
min_val <- min(values_pm25)
max_val <- max(values_pm25)
skew_val <- skewness(values_pm25)
percent_zeros <- sum(values_pm25 == 0)/length(values_pm25) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "PM5",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "PM25_Diagnosis_2000_201csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
```

```{r}
# Parameters
folder_path <- "D:/malaria"
output_path <- "D:/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (malar_et_2000.tif to malar_et_1tif)
files <- sprintf("malar_et_%02d.tif", 0:18)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_malaria <- rast(rasters)
# Pool all values into a single vector
values_malaria <- values(stack_malaria, mat = FALSE)
values_malaria <- values_malaria[!is.na(values_malaria)]
# Compute diagnostics
mean_val <- mean(values_malaria)
median_val <- median(values_malaria)
min_val <- min(values_malaria)
max_val <- max(values_malaria)
skew_val <- skewness(values_malaria)
percent_zeros <- sum(values_malaria == 0)/length(values_malaria) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "malaria",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "Malaria_Diagnosis_2000_201csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
```

```{r}
folder_path <- "D:/deareha"
output_path <- "D:/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (dear_et_2000.tif to dear_et_1tif)
files <- sprintf("dear_et_%02d.tif", 0:17)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_deareha <- rast(rasters)
# Pool all values into a single vector
values_deareha <- values(stack_deareha, mat = FALSE)
values_deareha <- values_deareha[!is.na(values_deareha)]
# Compute diagnostics
mean_val <- mean(values_deareha)
median_val <- median(values_deareha)
min_val <- min(values_deareha)
max_val <- max(values_deareha)
skew_val <- skewness(values_deareha)
percent_zeros <- sum(values_deareha == 0)/length(values_deareha) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "deareha",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "Dear_Diagnosis_2000_201csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
```

```{r}
# arameters
folder_path <- "D:/HIV"
output_path <- "D:/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (HIV_ETH_2000.tif to HIV_ETH_1tif)
files <- sprintf("HIV_ETH_%02d.tif", 0:19)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_hiv <- rast(rasters)
# Pool all values into a single vector
values_hiv <- values(stack_hiv, mat = FALSE)
values_hiv <- values_hiv[!is.na(values_hiv)]
# Compute diagnostics
mean_val <- mean(values_hiv)
median_val <- median(values_hiv)
min_val <- min(values_hiv)
max_val <- max(values_hiv)
skew_val <- skewness(values_hiv)
percent_zeros <- sum(values_hiv == 0)/length(values_hiv) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "HIV",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "HIV_Diagnosis_2000_201csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
```

```{r}
# Parameters
folder_path <- "D:/Sever wasting"
output_path <- "D:/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (SW 2000.tif to SW 201tif)
years <- 2000:2019
files <- sprintf("SW %d.tif", years)   # note the space instead of underscore
file_paths <- file.path(folder_path, files)
# Keep only existing files
existing_files <- file_paths[file.exists(file_paths)]
if(length(existing_files) == 0){
  stop("No raster files found in folder: ", folder_path)
}
# Load first layer safely
rasters <- lapply(existing_files, function(f){
  rast(f, lyrs = 1)
})
# Stack rasters
stack_sw <- rast(rasters)
# Pool all values into a single vector
values_sw <- values(stack_sw, mat = FALSE)
values_sw <- values_sw[!is.na(values_sw)]
# Compute diagnostics
mean_val <- mean(values_sw)
median_val <- median(values_sw)
min_val <- min(values_sw)
max_val <- max(values_sw)
skew_val <- skewness(values_sw)
percent_zeros <- sum(values_sw == 0)/length(values_sw) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "Severe Wasting",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "SevereWasting_Diagnosis_2000_201csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
```

```{r}
# Parameters
folder_path <- "D:/dpt3"
output_path <- "D:/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (DPT3_ET_2000.tif to DPT3_ET_1tif)
files <- sprintf("DPT3_ET_%02d.tif", 0:16)
file_paths <- file.path(folder_path, files)

# Keep only existing files
existing_files <- file_paths[file.exists(file_paths)]
if(length(existing_files) == 0){
  stop("No raster files found in folder: ", folder_path)
}
# Load first layer safely
rasters <- lapply(existing_files, function(f){
  rast(f, lyrs = 1)
})
# Stack rasters
stack_dpt3 <- rast(rasters)
# Pool all values into a single vector
values_dpt3 <- values(stack_dpt3, mat = FALSE)
values_dpt3 <- values_dpt3[!is.na(values_dpt3)]
# Compute diagnostics
mean_val <- mean(values_dpt3)
median_val <- median(values_dpt3)
min_val <- min(values_dpt3)
max_val <- max(values_dpt3)
skew_val <- skewness(values_dpt3)
percent_zeros <- sum(values_dpt3 == 0)/length(values_dpt3) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "DPT3",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "DPT3_Diagnosis_2000_201csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
```

```{r}
# Parameters
folder_path <- "D:/wash"
output_path <- "D:/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (wash 2000.tif to wash 201tif)
years <- 2000:2017
files <- sprintf("wash %d.tif", years)   # note the space
file_paths <- file.path(folder_path, files)
# Keep only existing files
existing_files <- file_paths[file.exists(file_paths)]
if(length(existing_files) == 0){
  stop("No raster files found in folder: ", folder_path)
}
# Load first layer safely
rasters <- lapply(existing_files, function(f){
  rast(f, lyrs = 1)
})
# Stack rasters
stack_wash <- rast(rasters)
# Pool all values into a single vector
values_wash <- values(stack_wash, mat = FALSE)
values_wash <- values_wash[!is.na(values_wash)]
# Compute diagnostics
mean_val <- mean(values_wash)
median_val <- median(values_wash)
min_val <- min(values_wash)
max_val <- max(values_wash)
skew_val <- skewness(values_wash)
percent_zeros <- sum(values_wash == 0)/length(values_wash) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "WASH",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "WASH_Diagnosis_2000_201csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
```

```{r}
# Parameters
folder_path <- "D:/water"
output_path <- "D:/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (water 2000.tif to water 201tif)
years <- 2000:2017
files <- sprintf("water %d.tif", years)   # note the space
file_paths <- file.path(folder_path, files)
# Keep only existing files
existing_files <- file_paths[file.exists(file_paths)]
if(length(existing_files) == 0){
  stop("No raster files found in folder: ", folder_path)
}
# Load first layer safely
rasters <- lapply(existing_files, function(f){
  rast(f, lyrs = 1)
})
# Stack rasters
stack_water <- rast(rasters)
# Pool all values into a single vector
values_water <- values(stack_water, mat = FALSE)
values_water <- values_water[!is.na(values_water)]
# Compute diagnostics
mean_val <- mean(values_water)
median_val <- median(values_water)
min_val <- min(values_water)
max_val <- max(values_water)
skew_val <- skewness(values_water)
percent_zeros <- sum(values_water == 0)/length(values_water) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "Water",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "Water_Diagnosis_2000_201csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
```

```{r}
# Parameters
folder_path <- "D:/education"
output_path <- "D:/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (Educ_2000.tif to Educ_201tif)
years <- 2000:2017
files <- sprintf("Educ_%d.tif", years)
file_paths <- file.path(folder_path, files)
# Keep only existing files
existing_files <- file_paths[file.exists(file_paths)]
if(length(existing_files) == 0){
  stop("No raster files found in folder: ", folder_path)
}
# Load first layer safely
rasters <- lapply(existing_files, function(f){
  rast(f, lyrs = 1)
})
# Stack rasters
stack_educ <- rast(rasters)
# Pool all values into a single vector
values_educ <- values(stack_educ, mat = FALSE)
values_educ <- values_educ[!is.na(values_educ)]
# Compute diagnostics
mean_val <- mean(values_educ)
median_val <- median(values_educ)
min_val <- min(values_educ)
max_val <- max(values_educ)
skew_val <- skewness(values_educ)
percent_zeros <- sum(values_educ == 0)/length(values_educ) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "Education",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "Education_Diagnosis_2000_201csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
```


####################################################################################################################

# COVARIATE PREPARATION PIPELINE FOR UNDER-FIVE MORTALITY PREDICTION

### Night time lights (NTL – Binary Urban/Rural

```{r}
#  LOAD ETHIOPIA BOUNDARY (for masking)
ethiopia_poly <- vect("D:/gadm41_ETH_0.shp")
ethiopia_poly_utm <- project(ethiopia_poly, crs(template))
stopifnot(crs(template) == crs(ethiopia_poly_utm))
```

```{r}
#  LOAD RAW NIGHT-TIME LIGHT RASTER (2000)
ntl_raw <- rast("D:/light 2000.tif")
if(nlyr(ntl_raw) > 1) ntl_raw <- ntl_raw[[1]]
#  PROJECT TO TEMPLATE CRS
ntl_proj <- if(!same.crs(ntl_raw, template)) project(ntl_raw, template) else ntl_raw
#  RESAMPLE TO 5 KM TEMPLATE
ntl_5km <- resample(ntl_proj, template, method="bilinear")
#  CROP + MASK TO ETHIOPIA
ntl_crop <- crop(ntl_5km, ethiopia_poly_utm)
ntl_clip <- mask(ntl_crop, ethiopia_poly_utm)
#  CREATE BINARY URBAN VARIABLE
# DN > 0 = 1 (urban)
# DN = 0  = 0 (rural)
ntl_binary <- classify(
  ntl_clip,
  rcl = matrix(c(-Inf, 0, 0,
                 0, Inf, 1),
               ncol=3, byrow=TRUE)
)
names(ntl_binary) <- "NTL_Urban_Binary"
#  FINAL CHECK
ntl_vals <- values(ntl_binary)
ntl_vals_nonNA <- ntl_vals[!is.na(ntl_vals)]
#  SAVE FINAL RASTER
output_raster <- "D:/ethiopia_nighttime_light_2000_5km_binary.tif"
writeRaster(
  ntl_binary,
  filename = output_raster,
  overwrite = TRUE,
  gdal = "COMPRESS=LZW"
)
#  METADATA
metadata <- list(
  variable_name = "Night-Time Lights (Binary Urban/Rural)",
  variable_code = "NTL00_5km_binary",
  year = 2000,
  country = "Ethiopia",
  source = "DMSP-OLS Night-Time Lights",
  original_units = "Digital Number (DN)",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(ntl_binary)),
  resolution_m = paste(res(ntl_binary), collapse=" x "),
  n_cells_total = ncell(ntl_binary),
  n_cells_valid = length(ntl_vals_nonNA),
  n_cells_missing = sum(is.na(ntl_vals)),
  classification_rule = "DN > 0 = 1 (urban), DN = 0 = 0 (rural)",
  proportion_urban = mean(ntl_vals_nonNA),
  proportion_rural = 1 - mean(ntl_vals_nonNA),
  unique_values = sort(unique(ntl_vals_nonNA)),
  transformation = "none (binary classification)",
  standardization = "none",
  imputation = "none",
  inla_ready = TRUE,
  spde_alignment_checked = TRUE
)
jsonlite::write_json(
  metadata,
  path = gsub(".tif$","_metadata.json",output_raster),
  pretty = TRUE,
  auto_unbox = TRUE
)
cat("Metadata JSON:", gsub(".tif$","_metadata.json",output_raster), "\n")
```


###  Dewpoint Temperature at 2 meters above ground level (K)  ERA5 Unit: Kelvin (K)

DEWPOINT TEMPERATURE COVARIATE PREPARATION
For Bayesian Hierarchical Model with INLA + SPDE
Target: Under-five mortality prediction at 5km resolution, Ethiopia 2000
  Source: ERA5 Reanalysis Original units: Kelvin (K)
Processing date: 2026-02-26

```{r}
#  LOAD RAW ERA5 DEWPOINT (Kelvin)
dew_raw <- rast("D:/dewPoi_2000.tif")
if (nlyr(dew_raw) > 1) dew_raw <- dew_raw[[1]]
#  PROJECT + RESAMPLE TO 5km GRID
dew_proj <- if (!same.crs(dew_raw, template)) project(dew_raw, template) else dew_raw
dew_5km <- resample(dew_proj, template, method = "bilinear")
#  CLIP TO ETHIOPIA
dew_crop <- crop(dew_5km, ethiopia_poly_utm)
dew_clip <- mask(dew_crop, ethiopia_poly_utm)
#  Z-SCORE STANDARDIZATION (Ethiopia pixels only)
vals <- values(dew_clip)
vals_eth <- vals[!is.na(vals)]
vals_z <- (vals - mean(vals_eth)) / sd(vals_eth)
values(dew_clip) <- vals_z
#  GEOMETRY CHECK
geom_check <- compareGeom(dew_clip, template, stopOnError = FALSE)
if (!geom_check) stop("Geometry mismatch with template.")
#  SAVE COVARIATE RASTER
output_filename <- "ethiopia_dewpoint_temp_2m_2000_5km_zscore.tif"
output_path <- file.path("D:/Adjusted_Raster", output_filename)
writeRaster(dew_clip, filename = output_path, overwrite = TRUE, gdal = "COMPRESS=LZW")
#  METADATA FOR COVARIATE STAGE ONLY
metadata <- list(
  variable_name = "Dewpoint Temperature at 2m",
  variable_code = "dewpoint",
  year = 2000,
  country = "Ethiopia",
  source = "ERA5 Reanalysis",
  source_doi = "24381/cds.adbb2d47",
  original_units = "Kelvin",
  covariate_processing = list(
    projection = "EPSG:32637 (UTM Zone 37N)",
    resampling = "Bilinear",
    masking = "Masked to Ethiopia national boundary",
    missing_handling = "NA values retained; Ethiopia pixels used for standardization",
    scaling = "Z-score standardization (mean=0, SD=1) based on Ethiopia pixels only"
  ),
  spatial_resolution_m = paste(round(res(dew_clip), 2), collapse = " x "),
  crs = as.character(crs(dew_clip)),
  n_cells = ncell(dew_clip),
  processing_steps = "projected, resampled (bilinear), masked to Ethiopia, standardized (z-score)",
  alignment_passed = geom_check,
  processing_date = as.character(Sys.Date())
)

# Save metadata
jsonlite::write_json(
  metadata, 
  path = gsub(".tif$", "_metadata.json", output_path),
  pretty = TRUE,
  auto_unbox = TRUE
)
```


# malaria prevalence / risk

```{r}
#  LOAD RAW MALARIA DATA
mal_raw <- rast("D:/malar_et_2000.tif")
if (nlyr(mal_raw) > 1) mal_raw <- mal_raw[[1]]
#  PROJECT + RESAMPLE TO 5km GRID
mal_proj <- if (!same.crs(mal_raw, template)) project(mal_raw, template) else mal_raw
mal_5km <- resample(mal_proj, template, method="bilinear")
#  CLIP TO ETHIOPIA
mal_crop <- crop(mal_5km, ethiopia_poly_utm)
mal_clip <- mask(mal_crop, ethiopia_poly_utm)
#  LOG1P TRANSFORMATION
vals <- values(mal_clip)
vals_eth <- vals[!is.na(vals)]
vals_log <- log1p(vals_eth)
#  Z-SCORE STANDARDIZATION
mean_val <- mean(vals_log)
sd_val <- sd(vals_log)
vals_std <- (vals_log - mean_val) / sd_val
# 6a. HANDLE EXTREME Z-SCORES
z_cap <- 5
n_extreme <- sum(abs(vals_std) > z_cap)
vals_std[vals_std > z_cap] <- z_cap
vals_std[vals_std < -z_cap] <- -z_cap
vals_full <- rep(NA, length(vals))
vals_full[!is.na(vals)] <- vals_std
values(mal_clip) <- vals_full
#  GEOMETRY CHECK
geom_check <- compareGeom(mal_clip, template, stopOnError = FALSE)
if (!geom_check) stop("Geometry mismatch with template.")
#  SAVE COVARIATE RASTER
output_filename <- "ethiopia_malaria_2000_5km_zlog.tif"
output_path <- file.path("D:/Adjusted_Raster", output_filename)
writeRaster(mal_clip, filename = output_path, overwrite = TRUE, gdal = "COMPRESS=LZW")
#  METADATA FOR COVARIATE STAGE ONLY
metadata <- list(
  variable_name = "Malaria",
  variable_code = "malaria",
  year = 2000,
  country = "Ethiopia",
  source = "Malaria Atlas Project",
  original_units = "continuous",
  covariate_processing = list(
    projection = "EPSG:32637 (UTM Zone 37N)",
    resampling = "Bilinear",
    masking = "Masked to Ethiopia national boundary",
    transformation = "log1p",
    scaling = "Z-score standardization (mean=0, SD=1, extremes capped at ±5)"
  ),
  spatial_resolution_m = paste(round(res(mal_clip), 2), collapse = " x "),
  crs = as.character(crs(mal_clip)),
  n_cells = ncell(mal_clip),
  n_cells_valid = sum(!is.na(values(mal_clip))),
  processing_steps = "projected, resampled (bilinear), masked to Ethiopia, log1p transformed, standardized (z-score), extreme z-scores capped at ±5",
  alignment_passed = geom_check,
  processing_date = as.character(Sys.Date())
)

jsonlite::write_json(
  metadata,
  path = gsub(".tif$", "_metadata.json", output_path),
  pretty = TRUE, auto_unbox = TRUE
)
```

# surface solar radiation downwards (likely in J/m² over a time period, based on ERA5)

```{r}
#  LOAD RAW SOLAR RADIATION
solar_raw <- rast("D:/rad_2000.tif")
if(nlyr(solar_raw) > 1) solar_raw <- solar_raw[[1]]
#  PROJECT + RESAMPLE TO 5KM GRID
solar_proj <- if(!same.crs(solar_raw, template)) project(solar_raw, template) else solar_raw
solar_5km <- resample(solar_proj, template, method="bilinear")
#  CROP + MASK TO ETHIOPIA
solar_crop <- crop(solar_5km, ethiopia_poly_utm)
solar_clip <- mask(solar_crop, ethiopia_poly_utm)
vals_before <- values(solar_clip)
vals_eth <- vals_before[!is.na(vals_before)]
#  Z-SCORE STANDARDIZATION
mean_val <- mean(vals_eth)
sd_val <- sd(vals_eth)
vals_std_eth <- (vals_eth - mean_val) / sd_val
vals_std_full <- rep(NA, length(vals_before))
vals_std_full[!is.na(vals_before)] <- vals_std_eth
values(solar_clip) <- vals_std_full
#  GEOMETRY CHECK
geom_check <- compareGeom(solar_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")
#  SAVE FINAL RASTER
output_filename <- "ethiopia_solar_radiation_2000_5km_zscore.tif"
output_path <- file.path("D:/Adjusted_Raster", output_filename)
writeRaster(solar_clip, filename=output_path, overwrite=TRUE, gdal="COMPRESS=LZW")
#  METADATA
metadata <- list(
  variable_name = "Surface Solar Radiation Downwards",
  variable_code = "solar",
  year = 2000,
  country = "Ethiopia",
  source = "ERA5",
  source_doi = "24381/cds.adbb2d47",
  original_units = "J/m²",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(solar_clip)),
  resolution_m = paste(res(solar_clip), collapse=" x "),
  n_cells = ncell(solar_clip),
  n_cells_valid = length(vals_eth),
  transformation = "none",
  standardization = "z-score",
  final_min = min(vals_eth),
  final_max = max(vals_eth),
  final_mean = mean(vals_eth),
  final_sd = sd(vals_eth),
  final_skewness = e1071::skewness(vals_eth),
  final_kurtosis = e1071::kurtosis(vals_eth),
  alignment_passed = geom_check,
  missing_final = sum(is.na(vals_before))
)
jsonlite::write_json(metadata, path=gsub(".tif$","_metadata.json",output_path), pretty=TRUE, auto_unbox=TRUE)
```


#  severe wasting prevalence at a certain spatial resolution

```{r}
#  LOAD RAW SEVERE WASTING RASTER
SW_raw <- rast("/Sever wasting/SW 2000.tif")
if(nlyr(SW_raw) > 1) SW_raw <- SW_raw[[1]]
#  LOG1P TRANSFORMATION
vals <- values(SW_raw)
vals_eth <- vals[!is.na(vals)]
vals_log <- log1p(vals_eth)
#  Z-SCORE STANDARDIZATION
mean_val <- mean(vals_log)
sd_val <- sd(vals_log)
vals_std <- (vals_log - mean_val) / sd_val

vals_full <- rep(NA, length(vals))
vals_full[!is.na(vals)] <- vals_std
values(SW_raw) <- vals_full
#  PROJECT + RESAMPLE TO 5KM TEMPLATE
SW_proj <- if(!same.crs(SW_raw, template)) project(SW_raw, template) else SW_raw
SW_5km <- resample(SW_proj, template, method="bilinear")
#  CROP + MASK TO ETHIOPIA
SW_crop <- crop(SW_5km, ethiopia_poly_utm)
SW_clip <- mask(SW_crop, ethiopia_poly_utm)
cat("Clipped to Ethiopia boundary, valid pixels:", sum(!is.na(values(SW_clip))), "\n")
#  GEOMETRY CHECK
geom_check <- compareGeom(SW_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")
#  SAVE FINAL RASTER
output_raster <- "D:/Adjusted_Raster/ethiopia_severe_wasting_2000_5km_zlog.tif"
writeRaster(SW_clip, filename=output_raster, overwrite=TRUE, gdal="COMPRESS=LZW")
#  METADATA
vals_final <- values(SW_clip)[!is.na(values(SW_clip))]
metadata <- list(
  variable_name = "Severe Wasting Prevalence",
  variable_code = "SW00_5km",
  year = 2000,
  country = "Ethiopia",
  source = "GBD Raster Data",
  original_units = "%",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(SW_clip)),
  resolution_m = paste(res(SW_clip), collapse=" x "),
  n_cells = ncell(SW_clip),
  n_cells_valid = length(vals_final),
  transformation = "log1p",
  standardization = "z-score",
  final_min = min(vals_final),
  final_max = max(vals_final),
  final_mean = mean(vals_final),
  final_sd = sd(vals_final),
  final_skewness = skewness(vals_final),
  alignment_passed = geom_check,
  missing_final = sum(is.na(values(SW_clip)))
)
write_json(metadata, path=gsub(".tif$","_metadata.json",output_raster), pretty=TRUE, auto_unbox=TRUE)
```


###   Water, Sanitation, and Hygiene (WASH, 2000)

```{r}
#  LOAD RAW WASH RASTER
Wash_raw <- rast("D:/wash 2000.tif")
if(nlyr(Wash_raw) > 1) Wash_raw <- Wash_raw[[1]]
#  LOG1P TRANSFORMATION
vals <- values(Wash_raw)
vals_nonNA <- vals[!is.na(vals)]
vals_log <- log1p(vals_nonNA)
#  Z-SCORE STANDARDIZATION
mean_val <- mean(vals_log)
sd_val <- sd(vals_log)
vals_std <- (vals_log - mean_val) / sd_val

vals_full <- rep(NA, length(vals))
vals_full[!is.na(vals)] <- vals_std
values(Wash_raw) <- vals_full
#  PROJECT + RESAMPLE TO 5KM TEMPLATE
Wash_proj <- if(!same.crs(Wash_raw, template)) project(Wash_raw, template) else Wash_raw
Wash_5km <- resample(Wash_proj, template, method="bilinear")
#  CROP + MASK TO ETHIOPIA
Wash_crop <- crop(Wash_5km, template)
Wash_clip <- mask(Wash_crop, template)
cat("Clipped to Ethiopia boundary, valid pixels:", sum(!is.na(values(Wash_clip))), "\n")
#  GEOMETRY CHECK
geom_check <- compareGeom(Wash_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")
#  SAVE FINAL RASTER
output_raster <- "D:/Adjusted_Raster/ethiopia_wash_2000_5km_zlog.tif"
writeRaster(Wash_clip, filename=output_raster, overwrite=TRUE, gdal="COMPRESS=LZW")
#  METADATA
vals_final <- values(Wash_clip)[!is.na(values(Wash_clip))]
metadata <- list(
  variable_name = "Water, Sanitation, and Hygiene (WASH)",
  variable_code = "Wash00_5km",
  year = 2000,
  country = "Ethiopia",
  source = "GBD Raster Data",
  original_units = "%",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(Wash_clip)),
  resolution_m = paste(res(Wash_clip), collapse=" x "),
  n_cells = ncell(Wash_clip),
  n_cells_valid = length(vals_final),
  transformation = "log1p",
  standardization = "z-score",
  final_min = min(vals_final),
  final_max = max(vals_final),
  final_mean = mean(vals_final),
  final_sd = sd(vals_final),
  final_skewness = skewness(vals_final),
  alignment_passed = geom_check,
  missing_final = sum(is.na(values(Wash_clip)))
)
write_json(metadata, path=gsub(".tif$","_metadata.json",output_raster), pretty=TRUE, auto_unbox=TRUE)
```


##  Education Coverage (EDUC, 2000)

```{r}
#  LOAD RAW EDUCATION RASTER
educ_raw <- rast("D:/Educ_2000.tif")
if(nlyr(educ_raw) > 1) educ_raw <- educ_raw[[1]]
#  Z-SCORE STANDARDIZATION (no transformation)
vals <- values(educ_raw)
vals_nonNA <- vals[!is.na(vals)]
mean_val <- mean(vals_nonNA)
sd_val <- sd(vals_nonNA)
vals_std <- (vals_nonNA - mean_val) / sd_val
# 3a. HANDLE EXTREME Z-SCORES
z_cap <- 5
n_extreme <- sum(abs(vals_std) > z_cap)
vals_std[vals_std > z_cap] <- z_cap
vals_std[vals_std < -z_cap] <- -z_cap
vals_full <- rep(NA, length(vals))
vals_full[!is.na(vals)] <- vals_std
values(educ_raw) <- vals_full
#  PROJECT + RESAMPLE TO 5KM TEMPLATE
educ_proj <- if(!same.crs(educ_raw, template)) project(educ_raw, template) else educ_raw
educ_5km <- resample(educ_proj, template, method="bilinear")
#  CROP + MASK TO ETHIOPIA
educ_crop <- crop(educ_5km, template)
educ_clip <- mask(educ_crop, template)
cat("Clipped to Ethiopia boundary, valid pixels:", sum(!is.na(values(educ_clip))), "\n")
#  GEOMETRY CHECK
geom_check <- compareGeom(educ_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")
#  SAVE FINAL RASTER
output_raster <- "D:/Adjusted_Raster/ethiopia_education_2000_5km_zscore.tif"
writeRaster(educ_clip, filename=output_raster, overwrite=TRUE, gdal="COMPRESS=LZW")
#  METADATA
vals_final <- values(educ_clip)[!is.na(values(educ_clip))]
metadata <- list(
  variable_name = "Education Coverage",
  variable_code = "educ00_5km",
  year = 2000,
  country = "Ethiopia",
  source = "GBD Raster Data",
  original_units = "%",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(educ_clip)),
  resolution_m = paste(res(educ_clip), collapse=" x "),
  n_cells = ncell(educ_clip),
  n_cells_valid = length(vals_final),
  transformation = "none",
  standardization = "z-score (extremes capped ±5)",
  final_min = min(vals_final),
  final_max = max(vals_final),
  final_mean = mean(vals_final),
  final_sd = sd(vals_final),
  final_skewness = skewness(vals_final),
  alignment_passed = geom_check,
  missing_final = sum(is.na(values(educ_clip)))
)
write_json(metadata, path=gsub(".tif$","_metadata.json",output_raster), pretty=TRUE, auto_unbox=TRUE)
```


##  DPT3 vaccination coverage

```{r}
#  LOAD RAW DPT3 RASTER
dpt3_raw <- rast("D:/DPT3_ET_2000.tif")
if(nlyr(dpt3_raw) > 1) dpt3_raw <- dpt3_raw[[1]]
#  Z-SCORE STANDARDIZATION (no transformation)
vals <- values(dpt3_raw)
vals_nonNA <- vals[!is.na(vals)]
mean_val <- mean(vals_nonNA)
sd_val <- sd(vals_nonNA)
vals_std <- (vals_nonNA - mean_val) / sd_val
# 3a. HANDLE EXTREME Z-SCORES (cap at ±5)
z_cap <- 5
n_extreme <- sum(abs(vals_std) > z_cap)
vals_std[vals_std > z_cap] <- z_cap
vals_std[vals_std < -z_cap] <- -z_cap
vals_full <- rep(NA, length(vals))
vals_full[!is.na(vals)] <- vals_std
values(dpt3_raw) <- vals_full
#  PROJECT + RESAMPLE TO 5KM TEMPLATE
dpt3_proj <- if(!same.crs(dpt3_raw, template)) project(dpt3_raw, template) else dpt3_raw
dpt3_5km <- resample(dpt3_proj, template, method="bilinear")
#  CROP + MASK TO ETHIOPIA
dpt3_crop <- crop(dpt3_5km, template)
dpt3_clip <- mask(dpt3_crop, template)
#  GEOMETRY CHECK
geom_check <- compareGeom(dpt3_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")
#  SAVE FINAL RASTER
output_raster <- "D:/Adjusted_Raster/ethiopia_dpt3_2000_5km_zscore.tif"
writeRaster(dpt3_clip, filename=output_raster, overwrite=TRUE, gdal="COMPRESS=LZW")
#  METADATA
vals_final <- values(dpt3_clip)[!is.na(values(dpt3_clip))]
metadata <- list(
  variable_name = "DPT3 Vaccination Coverage",
  variable_code = "dpt300_5km",
  year = 2000,
  country = "Ethiopia",
  source = "GBD Raster Data",
  original_units = "%",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(dpt3_clip)),
  resolution_m = paste(res(dpt3_clip), collapse=" x "),
  n_cells = ncell(dpt3_clip),
  n_cells_valid = length(vals_final),
  transformation = "none",
  standardization = "z-score (extremes capped ±5)",
  final_min = min(vals_final),
  final_max = max(vals_final),
  final_mean = mean(vals_final),
  final_sd = sd(vals_final),
  final_skewness = skewness(vals_final),
  alignment_passed = geom_check,
  missing_final = sum(is.na(values(dpt3_clip)))
)
write_json(metadata, path=gsub(".tif$","_metadata.json",output_raster), pretty=TRUE, auto_unbox=TRUE)
```

###   Improved Water Coverage

```{r}
#  LOAD RAW IMPROVED WATER RASTER
Wate_raw <- rast("D:/water 2000.tif")
if(nlyr(Wate_raw) > 1) Wate_raw <- Wate_raw[[1]]
#  Z-SCORE STANDARDIZATION (no transformation, no capping)
vals <- values(Wate_raw)
vals_nonNA <- vals[!is.na(vals)]
mean_val <- mean(vals_nonNA)
sd_val <- sd(vals_nonNA)
vals_std <- (vals_nonNA - mean_val) / sd_val
vals_full <- rep(NA, length(vals))
vals_full[!is.na(vals)] <- vals_std
values(Wate_raw) <- vals_full
#  PROJECT + RESAMPLE TO 5KM TEMPLATE
Wate_proj <- if(!same.crs(Wate_raw, template)) project(Wate_raw, template) else Wate_raw
Wate_5km <- resample(Wate_proj, template, method="bilinear")
#  CROP + MASK TO ETHIOPIA
Wate_crop <- crop(Wate_5km, template)
Wate_clip <- mask(Wate_crop, template)
#  GEOMETRY CHECK
geom_check <- compareGeom(Wate_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")
#  SAVE FINAL RASTER
output_raster <- "D:/Adjusted_Raster/ethiopia_water_2000_5km_zscore.tif"
writeRaster(Wate_clip, filename=output_raster, overwrite=TRUE, gdal="COMPRESS=LZW")
cat("Raster saved:", output_raster, "\n")

#  METADATA
vals_final <- values(Wate_clip)[!is.na(values(Wate_clip))]
metadata <- list(
  variable_name = "Improved Water Coverage",
  variable_code = "Wate00_5km",
  year = 2000,
  country = "Ethiopia",
  source = "GBD Raster Data",
  original_units = "%",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(Wate_clip)),
  resolution_m = paste(res(Wate_clip), collapse=" x "),
  n_cells = ncell(Wate_clip),
  n_cells_valid = length(vals_final),
  transformation = "none",
  standardization = "z-score",
  final_min = min(vals_final),
  final_max = max(vals_final),
  final_mean = mean(vals_final),
  final_sd = sd(vals_final),
  final_skewness = skewness(vals_final),
  alignment_passed = geom_check,
  missing_final = sum(is.na(values(Wate_clip)))
)
write_json(metadata, path=gsub(".tif$","_metadata.json",output_raster), pretty=TRUE, auto_unbox=TRUE)
```


##  diarrhea incidence among under-five children

```{r}
#  LOAD RAW DIARRHEA INCIDENCE RASTER
dear_raw <- rast("D:/dear_et_2000.tif")
if(nlyr(dear_raw) > 1) dear_raw <- dear_raw[[1]]
cat("Raw Diarrhea Incidence raster loaded\n")

#  Z-SCORE STANDARDIZATION (no transformation, no capping)
vals <- values(dear_raw)
vals_nonNA <- vals[!is.na(vals)]
mean_val <- mean(vals_nonNA)
sd_val <- sd(vals_nonNA)
vals_std <- (vals_nonNA - mean_val) / sd_val

vals_full <- rep(NA, length(vals))
vals_full[!is.na(vals)] <- vals_std
values(dear_raw) <- vals_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1)\n")

#  PROJECT + RESAMPLE TO 5KM TEMPLATE
dear_proj <- if(!same.crs(dear_raw, template)) project(dear_raw, template) else dear_raw
dear_5km <- resample(dear_proj, template, method="bilinear")
cat("Projected and resampled to 5km\n")

#  CROP + MASK TO ETHIOPIA
dear_crop <- crop(dear_5km, template)
dear_clip <- mask(dear_crop, template)
cat("Clipped to Ethiopia boundary, valid pixels:", sum(!is.na(values(dear_clip))), "\n")

#  GEOMETRY CHECK
geom_check <- compareGeom(dear_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")

#  SAVE FINAL RASTER
output_raster <- "D:/Adjusted_Raster/ethiopia_deareaha_2000_5km_zscore.tif"
writeRaster(dear_clip, filename=output_raster, overwrite=TRUE, gdal="COMPRESS=LZW")
cat("Raster saved:", output_raster, "\n")

#  METADATA
vals_final <- values(dear_clip)[!is.na(values(dear_clip))]
metadata <- list(
  variable_name = "Diarrhea Incidence Under-5",
  variable_code = "dear00_5km",
  year = 2000,
  country = "Ethiopia",
  source = "GBD Raster Data",
  original_units = "cases per 1000 children",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(dear_clip)),
  resolution_m = paste(res(dear_clip), collapse=" x "),
  n_cells = ncell(dear_clip),
  n_cells_valid = length(vals_final),
  transformation = "none",
  standardization = "z-score",
  final_min = min(vals_final),
  final_max = max(vals_final),
  final_mean = mean(vals_final),
  final_sd = sd(vals_final),
  final_skewness = skewness(vals_final),
  alignment_passed = geom_check,
  missing_final = sum(is.na(values(dear_clip)))
)
write_json(metadata, path=gsub(".tif$","_metadata.json",output_raster), pretty=TRUE, auto_unbox=TRUE)
```



##  Normalized Difference Vegetation Index

```{r}
#  LOAD RAW NDVI RASTER
Ndvi_raw <- rast("D:/NDVI_2000_Ethiopia.tif")
if(nlyr(Ndvi_raw) > 1) Ndvi_raw <- Ndvi_raw[[1]]
cat("Raw NDVI raster loaded\n")
cat("Value range:", range(values(Ndvi_raw), na.rm=TRUE), 
    "Mean:", mean(values(Ndvi_raw), na.rm=TRUE),
    "SD:", sd(values(Ndvi_raw), na.rm=TRUE), "\n\n")
#  Z-SCORE STANDARDIZATION (no transformation, no capping)
vals <- values(Ndvi_raw)
vals_nonNA <- vals[!is.na(vals)]
mean_val <- mean(vals_nonNA)
sd_val <- sd(vals_nonNA)
vals_std <- (vals_nonNA - mean_val) / sd_val
vals_full <- rep(NA, length(vals))
vals_full[!is.na(vals)] <- vals_std
values(Ndvi_raw) <- vals_full
#  PROJECT + RESAMPLE TO 5KM TEMPLATE
Ndvi_proj <- if(!same.crs(Ndvi_raw, template)) project(Ndvi_raw, template) else Ndvi_raw
Ndvi_5km <- resample(Ndvi_proj, template, method="bilinear")
#  CROP + MASK TO ETHIOPIA
Ndvi_crop <- crop(Ndvi_5km, template)
Ndvi_clip <- mask(Ndvi_crop, template)
cat("Clipped to Ethiopia boundary, valid pixels:", sum(!is.na(values(Ndvi_clip))), "\n")
#  IMPUTE MISSING VALUES (0 in z-space)
vals_clip <- values(Ndvi_clip)
values(Ndvi_clip) <- vals_clip
cat("Missing pixels after imputation:", sum(is.na(values(Ndvi_clip))), "\n")
#  SAVE FINAL RASTER
output_raster <- "D:/Adjusted_Raster/ethiopia_ndvi_2000_5km_zscore.tif"
writeRaster(Ndvi_clip, filename=output_raster, overwrite=TRUE, gdal="COMPRESS=LZW")
#  METADATA
vals_final <- values(Ndvi_clip)[!is.na(values(Ndvi_clip))]
metadata <- list(
  variable_name = "NDVI",
  variable_code = "Ndvi00_5km",
  year = 2000,
  country = "Ethiopia",
  source = "NASA",
  original_units = "unitless (fraction)",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(Ndvi_clip)),
  resolution_m = paste(res(Ndvi_clip), collapse=" x "),
  n_cells = ncell(Ndvi_clip),
  n_cells_valid = length(vals_final),
  transformation = "none",
  standardization = "z-score",
  final_min = min(vals_final),
  final_max = max(vals_final),
  final_mean = mean(vals_final),
  final_sd = sd(vals_final),
  final_skewness = skewness(vals_final),
  missing_final = sum(is.na(values(Ndvi_clip)))
)
write_json(metadata, path=gsub(".tif$","_metadata.json",output_raster), pretty=TRUE, auto_unbox=TRUE)
cat("Covariate-stage metadata saved\n")
#  DIAGNOSTICS TABLE
diagnostic_table <- data.frame(
  variable = "Ndvi00_5km",
  min = min(vals_final),
  max = max(vals_final),
  range = max(vals_final) - min(vals_final),
  mean = mean(vals_final),
  median = median(vals_final),
  sd = sd(vals_final),
  n_missing = sum(is.na(values(Ndvi_clip))),
  n_non_missing = length(vals_final),
  prop_missing = sum(is.na(values(Ndvi_clip))) / ncell(Ndvi_clip),
  skewness = skewness(vals_final),
  suggested_transformation = "none",
  suggested_rescaling = "Z-score",
  imputation = "Mean in z-score space (0)"
)
write_xlsx(diagnostic_table, "/data managment/diagnose/NDVI00_5km.xlsx")
write.csv(diagnostic_table, "/data managment/diagnose/NDVI00_5km.csv", row.names=FALSE)
cat("Raster:", output_raster, "\n")
cat("Metadata JSON:", gsub(".tif$","_metadata.json",output_raster), "\n")
```

##  Fine Particulate Matter (PM25)

```{r}
#  LOAD RAW PM5 RASTER
raster_path <- "D:/pm25_et_2000.tif"
raster_name <- "PM5_2000"
r <- rast(raster_path)
if(nlyr(r) > 1) r <- r[[1]]
cat("Raw PM5 raster loaded\n")
cat("Value range:", range(values(r), na.rm=TRUE), 
    "Mean:", mean(values(r), na.rm=TRUE),
    "SD:", sd(values(r), na.rm=TRUE), "\n\n")
#  Z-SCORE STANDARDIZATION (no transformation, no capping)
vals <- values(r)
vals_nonNA <- vals[!is.na(vals)]
vals_std <- (vals_nonNA - mean(vals_nonNA)) / sd(vals_nonNA)
vals_full <- rep(NA, length(vals))
vals_full[!is.na(vals)] <- vals_std
values(r) <- vals_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1)\n")
#  PROJECT + RESAMPLE TO 5KM TEMPLATE
r_proj <- if(!same.crs(r, template)) project(r, template) else r
r_5km <- resample(r_proj, template, method="bilinear")
cat("Projected and resampled to 5km\n")
#  CROP + MASK TO ETHIOPIA
r_crop <- crop(r_5km, template)
r_clip <- mask(r_crop, template)
cat("Clipped to Ethiopia boundary, valid pixels:", sum(!is.na(values(r_clip))), "\n")
#  IMPUTE MISSING VALUES (0 in z-space)
vals_clip <- values(r_clip)
values(r_clip) <- vals_clip
cat("Missing pixels after imputation:", sum(is.na(values(r_clip))), "\n")
#  SAVE FINAL RASTER
output_raster <- paste0("D:/Adjusted_Raster/", raster_name, "_5km.tif")
writeRaster(r_clip, filename=output_raster, overwrite=TRUE, gdal="COMPRESS=LZW")
cat("Raster saved:", output_raster, "\n")
#  METADATA
vals_final <- values(r_clip)[!is.na(values(r_clip))]
metadata <- list(
  variable_name = "PM5",
  variable_code = "PM25_2000_5km",
  year = 2000,
  country = "Ethiopia",
  source = "GBD",
  original_units = "µg/m³",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(r_clip)),
  resolution_m = paste(res(r_clip), collapse=" x "),
  n_cells = ncell(r_clip),
  n_cells_valid = length(vals_final),
  transformation = "none",
  standardization = "z-score",
  final_min = min(vals_final),
  final_max = max(vals_final),
  final_mean = mean(vals_final),
  final_sd = sd(vals_final),
  final_skewness = skewness(vals_final),
  missing_final = sum(is.na(values(r_clip)))
)
jsonlite::write_json(metadata, path=gsub(".tif$","_metadata.json",output_raster), pretty=TRUE, auto_unbox=TRUE)
cat("Raster:", output_raster, "\n")
cat("Metadata JSON:", gsub(".tif$","_metadata.json",output_raster), "\n")
```

##  HIV Prevalence

```{r}
#  Load raw HIV raster
HIV_2000 <- rast("/HIV/HIV_ETH_2000.tif")
plot(HIV_2000, main="HIV Prevalence 2000 (Raw)")
summary(HIV_2000[[1]])
cat("Missing pixels:", sum(is.na(values(HIV_2000[[1]]))), "\n\n")
#  Apply log1p transformation
hiv_trans <- log1p(values(HIV_2000[[1]]))  
#  Z-score standardization
hiv_std <- (hiv_trans - mean(hiv_trans, na.rm=TRUE)) / sd(hiv_trans, na.rm=TRUE)
values(HIV_2000[[1]]) <- hiv_std
summary(HIV_2000)
hist(hiv_std, breaks=50, main="Distribution of HIV Prevalence (Z-score)", xlab="Z-score")
#  Project to template CRS
HIV_utm <- project(HIV_2000, crs(template))
#  Resample to 5 km template
HIV_5km <- resample(HIV_utm[[1]], template, method="bilinear")
#  Crop and mask to Ethiopia boundary
HIV_5km_crop <- crop(HIV_5km, template)
HIV_5km_clip <- mask(HIV_5km_crop, template)
plot(HIV_5km_clip, main="HIV Prevalence 2000 (Z-score, Clipped)")
summary(HIV_5km_clip)
cat("Missing pixels after clipping:", sum(is.na(values(HIV_5km_clip))), "\n\n")
#  Metadata
hiv_values <- values(HIV_5km_clip)
hiv_values_nonNA <- hiv_values[!is.na(hiv_values)]
HIV_metadata <- list(
  variable = "HIV Prevalence",
  year = 2000,
  source = "GBD",
  units = "fraction (0-1)",
  spatial_resolution_km = 5,
  crs = crs(HIV_5km_clip),
  n_rows = nrow(HIV_5km_clip),
  n_cols = ncol(HIV_5km_clip),
  n_cells_total = ncell(HIV_5km_clip),
  n_cells_valid = sum(!is.na(values(HIV_5km_clip))),
  n_cells_missing = sum(is.na(values(HIV_5km_clip))),
  min = min(hiv_values_nonNA),
  max = max(hiv_values_nonNA),
  mean = mean(hiv_values_nonNA),
  sd = sd(hiv_values_nonNA),
  skewness = skewness(hiv_values_nonNA),
  transformation = "log1p",
  standardization = "z-score",
  processing_notes = "Cropped, masked, Z-score standardized, missing values imputed with 0"
)
jsonlite::write_json(
  HIV_metadata, 
  "D:/Adjusted_Raster/HIV00_5km_metadata.json",
  pretty = TRUE
)
cat("HIV metadata saved\n\n")
#  Save final raster
writeRaster(
  HIV_5km_clip,
  "D:/Adjusted_Raster/ethiopia_hiv_2000_5km_zscore.tif",
  overwrite = TRUE
)
```


###  Potential Evaporation (EVAP)

```{r}
raw_raster_path <- "R:/evap_2000.tif"
output_dir <- "D:/Adjusted_Raster"
# Create output folder if missing
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
  cat("Output directory created\n")
} else {
  cat("Output directory already exists\n")
}
#  LOAD RAW ERA5 POTENTIAL EVAPORATION
evap_raw <- rast(raw_raster_path)
if (nlyr(evap_raw) > 1) evap_raw <- evap_raw[[1]]
#  LOG TRANSFORMATION (log1p)
vals_raw <- values(evap_raw)
vals_log <- log1p(vals_raw)
values(evap_raw) <- vals_log
#  PROJECT + RESAMPLE TO 5km GRID
evap_proj <- if (!same.crs(evap_raw, template)) {
  project(evap_raw, template)
} else {
  evap_raw
}
evap_5km <- resample(evap_proj, template, method = "bilinear")
#  CLIP TO ETHIOPIA
evap_crop <- crop(evap_5km, ethiopia_poly_utm)
evap_clip <- mask(evap_crop, ethiopia_poly_utm)
#  Z-SCORE STANDARDIZATION (Ethiopia pixels only)
vals <- values(evap_clip)
vals_eth <- vals[!is.na(vals)]
vals_z <- (vals - mean(vals_eth)) / sd(vals_eth)
# Cap extreme values (lower tail ≈ -8)
z_cap <- 5
vals_z[vals_z > 5]  <- 5
vals_z[vals_z < -5] <- -5
values(evap_clip) <- vals_z
cat("Standardization complete (mean≈0, sd≈1, capped ±5)\n")
#  GEOMETRY CHECK
geom_check <- compareGeom(evap_clip, template, stopOnError = FALSE)
if (!geom_check) {
  stop("Geometry mismatch detected.")
} else {
  cat("Geometry alignment: PASS\n")
}
#  SAVE FINAL RASTER
output_raster_path <- file.path(
  output_dir,
  "ethiopia_potential_evaporation_2000_5km_zscore.tif"
)
writeRaster(
  evap_clip,
  filename = output_raster_path,
  overwrite = TRUE,
  gdal = "COMPRESS=LZW"
)
# 1 GENERATE METADATA
evap_values <- values(evap_clip)
evap_values_nonNA <- evap_values[!is.na(evap_values)]
EVAP_metadata <- list(
  variable_name = "Potential Evaporation",
  variable_code = "potential_evaporation",
  year = 2000,
  country = "Ethiopia",
  source = "ERA5 Reanalysis",
  source_doi = "24381/cds.adbb2d47",
  original_units = "mm/day",
  covariate_processing = list(
    transformation = "log1p",
    projection = as.character(crs(evap_clip)),
    resampling = "Bilinear",
    masking = "Masked to Ethiopia national boundary",
    scaling = "Z-score standardization (mean=0, SD=1) with ±5 cap"
  ),
  spatial_resolution_m = paste(round(res(evap_clip), 2), collapse = " x "),
  n_cells_total = ncell(evap_clip),
  n_cells_valid = sum(!is.na(evap_values)),
  n_cells_missing = sum(is.na(evap_values)),
  min = min(evap_values_nonNA),
  max = max(evap_values_nonNA),
  mean = mean(evap_values_nonNA),
  sd = sd(evap_values_nonNA),
  alignment_passed = geom_check,
  processing_date = as.character(Sys.Date())
)

metadata_path <- file.path(
  output_dir,
  "eth_2000_potential_evaporation_5km_z_metadata.json"
)
write_json(
  EVAP_metadata,
  path = metadata_path,
  pretty = TRUE,
  auto_unbox = TRUE
)
```

##  Runoff

```{r}
#  Load raw runoff raster
run_raw <- rast("D:/run_2000.tif")
if (nlyr(run_raw) > 1) run_raw <- run_raw[[1]]
run_vals <- values(run_raw)
cat("Raw runoff: missing pixels =", sum(is.na(run_vals)), "\n")
cat("Value range:", range(run_vals, na.rm=TRUE), "\n\n")
#  Apply log1p transformation
run_trans <- log1p(run_vals)
#  Z-score standardization
mean_val <- mean(run_trans, na.rm=TRUE)
sd_val   <- sd(run_trans, na.rm=TRUE)
run_std <- (run_trans - mean_val) / sd_val
#  Handle extreme Z-scores by capping at ±5
run_std[run_std > 5] <- 5
run_std[run_std < -5] <- -5
values(run_raw) <- run_std
# Quick diagnostic plot
hist(run_std, breaks=50, main="Runoff (Z-score capped ±5)", xlab="Z-score")
#  Project to template CRS
run_proj <- project(run_raw, template)
#  Resample to 5 km template
run_5km <- resample(run_proj, template, method="bilinear")
#  Crop and mask to Ethiopia boundary
run_crop <- crop(run_5km, ethiopia_poly_utm)
run_clip <- mask(run_crop, ethiopia_poly_utm)
plot(run_clip, main="Standardized Runoff 2000 (5 km, Clipped, Z capped)")
#  Create metadata
run_final_vals <- values(run_clip)
run_nonNA <- run_final_vals[!is.na(run_final_vals)]
Runoff_metadata <- list(
  variable = "Runoff",
  year = 2000,
  source = "ERA5",
  units = "m³/s or mm/day",
  spatial_resolution_km = 5,
  crs = crs(run_clip),
  n_rows = nrow(run_clip),
  n_cols = ncol(run_clip),
  n_cells_total = ncell(run_clip),
  n_cells_valid = sum(!is.na(values(run_clip))),
  n_cells_missing = sum(is.na(values(run_clip))),
  min = min(run_nonNA),
  max = max(run_nonNA),
  mean = mean(run_nonNA),
  sd = sd(run_nonNA),
  skewness = skewness(run_nonNA),
  transformation = "log1p",
  standardization = "z-score",
  extreme_handling = "Z-scores capped at ±5",
  processing_notes = "Cropped, masked, Z-score standardized, log1p transformed"
)
jsonlite::write_json(
  Runoff_metadata, 
  "D:/Adjusted_Raster/Runoff00_5km_metadata.json",
  pretty = TRUE
)
cat("Runoff metadata saved\n\n")
#  Save final raster
output_path <- "D:/Adjusted_Raster/ethiopia_runoff_2000_5km_zscore.tif"
writeRaster(run_clip, filename=output_path, overwrite=TRUE, gdal="COMPRESS=LZW")
```

##  Slope (constant)

```{r}
#  Load raw Slope raster
Slo_2000 <- rast("D:/eth_slope_merit103_100m_vtif")
plot(Slo_2000, main="Raw Slope (2000)")
summary(Slo_2000)
sum(is.na(values(Slo_2000[[1]])))  # check missing pixels
#  Log1p transformation (slope is strictly positive and right-skewed)
slo_trans <- log1p(values(Slo_2000[[1]]))  
#  Z-score standardization
slo_std <- (slo_trans - mean(slo_trans, na.rm = TRUE)) / sd(slo_trans, na.rm = TRUE)
#  Handle extreme Z-scores by capping at ±5
slo_std[slo_std > 5] <- 5
slo_std[slo_std < -5] <- -5
values(Slo_2000[[1]]) <- slo_std
# Quick diagnostic check
summary(Slo_2000)
hist(slo_std, breaks=50, main="Distribution of Slope (Z-score capped ±5)", xlab="Z-score")
#  Project to template CRS
Slo_utm <- project(Slo_2000, crs(template))
#  Resample to 5 km
Slo00_5km <- resample(Slo_utm[[1]], template, method="bilinear")
#  Crop and mask to template extent
Slo00_5km_crop <- crop(Slo00_5km, template)
Slo00_5km_clip <- mask(Slo00_5km_crop, template)
plot(Slo00_5km_clip, main="Standardized Slope (5 km, Clipped)")
summary(Slo00_5km_clip)
sum(is.na(values(Slo00_5km_clip)))  # check missing pixels
#  Create metadata
slo_values <- values(Slo00_5km_clip)
slo_nonNA <- slo_values[!is.na(slo_values)]
Slope_metadata <- list(
  variable = "Slope",   
  year = 2000,
  units = "degrees",
  source = "Digital Elevation Models (DEMs)",
  spatial_resolution_km = 5,
  crs = crs(Slo00_5km_clip),
  n_rows = nrow(Slo00_5km_clip),
  n_cols = ncol(Slo00_5km_clip),
  n_cells_total = ncell(Slo00_5km_clip),
  n_cells_valid = sum(!is.na(values(Slo00_5km_clip))),
  n_cells_missing = sum(is.na(values(Slo00_5km_clip))),
  min = min(slo_nonNA),
  max = max(slo_nonNA),
  mean = mean(slo_nonNA),
  sd = sd(slo_nonNA),
  skewness = skewness(slo_nonNA),
  transformation = "log1p",
  standardization = "z-score",
  extreme_handling = "Z-scores capped at ±5",
  processing_notes = "Cropped, masked, Z-score standardized, log1p transformed"
)
jsonlite::write_json(
  Slope_metadata, 
  "D:/Adjusted_Raster/Slo00_5km_metadata.json",
  pretty = TRUE
)
cat("Slope metadata saved\n")
#  Save final raster
writeRaster(
  Slo00_5km_clip,
  "D:/Adjusted_Raster/ethiopia_slope_5km_zscore.tif",
  overwrite = TRUE
)
```

##  Exclusive Breastfeeding

```{r}
#  LOAD RAW EBF RASTER (adjust year as needed)
#    Example: year 2000 (choose any from 2000-2019)
ebf_raw <- rast("D:/EBF 200tif")
if(nlyr(ebf_raw) > 1) ebf_raw <- ebf_raw[[1]]
#  PROJECT + RESAMPLE TO 5KM GRID (use NEAREST to preserve zeros)
ebf_proj <- if(!same.crs(ebf_raw, template)) project(ebf_raw, template) else ebf_raw
ebf_5km <- resample(ebf_proj, template, method="near")
#  CROP + MASK TO ETHIOPIA
ebf_crop <- terra::crop(ebf_5km, ethiopia_poly_utm)
ebf_clip <- terra::mask(ebf_crop, ethiopia_poly_utm)
vals_before <- values(ebf_clip)
vals_eth <- vals_before[!is.na(vals_before)]
#  Z-SCORE STANDARDIZATION
mean_val <- mean(vals_eth)
sd_val <- sd(vals_eth)
vals_std_eth <- (vals_eth - mean_val) / sd_val
vals_std_full <- rep(NA, length(vals_before))
vals_std_full[!is.na(vals_before)] <- vals_std_eth
values(ebf_clip) <- vals_std_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1)\n")
#  GEOMETRY CHECK
geom_check <- compareGeom(ebf_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")
#  SAVE FINAL RASTER
output_filename <- "ethiopia_5km_ebf_2000_zscore.tif"
output_path <- file.path("D:/adjusted_covaraite", output_filename)
# writeRaster(ebf_clip, filename=output_path, overwrite=TRUE, gdal="COMPRESS=LZW")
# cat("Raster saved:", output_path, "\n")
#  METADATA
metadata <- list(
  variable_name = "Exclusive Breastfeeding (EBF) prevalence",
  variable_code = "EBF",
  year = 2000,                                   # adjust per file
  country = "Ethiopia",
  source = "DHS / UNICEF estimates",
  original_units = "percent",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(ebf_clip)),
  resolution_m = paste(res(ebf_clip), collapse=" x "),
  n_cells = ncell(ebf_clip),
  n_cells_valid = length(vals_eth),
  percent_zeros_original = 414349097,          # from diagnosis table
  skewness_original = 0.340602945,
  transformation = "none",
  standardization = "z-score",
  final_min = min(vals_eth),
  final_max = max(vals_eth),
  final_mean = mean(vals_eth),
  final_sd = sd(vals_eth),
  final_skewness = e1071::skewness(vals_eth),
  final_kurtosis = e1071::kurtosis(vals_eth),
  alignment_passed = geom_check,
  missing_final = sum(is.na(vals_before))
)
output_path1 <- file.path("D:/metadara/2000", output_filename)
# jsonlite::write_json(metadata, path=gsub(".tif$","_metadata.json",output_path1), 
 #                    pretty=TRUE, auto_unbox=TRUE)
```

##  CHIRPS 

```{r}
#  LOAD RAW CHIRPS RASTER (adjust path & year as needed)
#    Example: annual total precipitation for year 2000
chirps_raw <- rast("D:/U5M/Covaraite original/chrips/chirpsETH_2000.tif")
if(nlyr(chirps_raw) > 1) chirps_raw <- chirps_raw[[1]]   # take first layer if stack
#  PROJECT + RESAMPLE TO 5KM GRID
chirps_proj <- if(!same.crs(chirps_raw, template)) project(chirps_raw, template) else chirps_raw
chirps_5km <- resample(chirps_proj, template, method="bilinear")
#  CROP + MASK TO ETHIOPIA
chirps_crop <- terra::crop(chirps_5km, ethiopia_poly_utm)
chirps_clip <- terra::mask(chirps_crop, ethiopia_poly_utm)
vals_before <- values(chirps_clip)
vals_eth <- vals_before[!is.na(vals_before)]
#  Z-SCORE STANDARDIZATION
mean_val <- mean(vals_eth)
sd_val <- sd(vals_eth)
vals_std_eth <- (vals_eth - mean_val) / sd_val
vals_std_full <- rep(NA, length(vals_before))
vals_std_full[!is.na(vals_before)] <- vals_std_eth
values(chirps_clip) <- vals_std_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1)\n")

#  GEOMETRY CHECK
geom_check <- compareGeom(chirps_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")

#  SAVE FINAL RASTER
output_filename <- "ethiopia_5km_chirps_2000_zscore.tif"
output_path <- file.path("D:/Adjusted_Raster", output_filename)
# writeRaster(chirps_clip, filename=output_path, overwrite=TRUE, gdal="COMPRESS=LZW")
# cat("Raster saved:", output_path, "\n")

#  METADATA
metadata <- list(
  variable_name = "CHIRPS Precipitation (annual total)",
  variable_code = "chirps",
  year = 2000,                                   # change as needed
  country = "Ethiopia",
  source = "CHIRPS (Climate Hazards Group InfraRed Precipitation with Station data)",
  source_doi = "15780/G2RP4Q",
  original_units = "mm",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(chirps_clip)),
  resolution_m = paste(res(chirps_clip), collapse=" x "),
  n_cells = ncell(chirps_clip),
  n_cells_valid = length(vals_eth),
  percent_zeros_original = 435,                # from diagnosis table
  skewness_original = 114003565,
  transformation = "none",
  standardization = "z-score",
  final_min = min(vals_eth),
  final_max = max(vals_eth),
  final_mean = mean(vals_eth),
  final_sd = sd(vals_eth),
  final_skewness = e1071::skewness(vals_eth),
  final_kurtosis = e1071::kurtosis(vals_eth),
  alignment_passed = geom_check,
  missing_final = sum(is.na(vals_before))
)
```

## Elevation

```{r}
#  LOAD RAW ELEVATION RASTER (constant, no year)
elev_raw <- rast("D:/eth_elevation_merit103_100m_vtif")  # adjust filename as needed
if(nlyr(elev_raw) > 1) elev_raw <- elev_raw[[1]]
cat("Raw elevation raster loaded\n")

#  PROJECT + RESAMPLE TO 5KM GRID (use NEAREST to preserve exact elevations)
elev_proj <- if(!same.crs(elev_raw, template)) project(elev_raw, template) else elev_raw
elev_5km <- resample(elev_proj, template, method="near")
cat("Projected and resampled to 5km (nearest neighbor)\n")

#  CROP + MASK TO ETHIOPIA
elev_crop <- terra::crop(elev_5km, ethiopia_poly_utm)
elev_clip <- terra::mask(elev_crop, ethiopia_poly_utm)
vals_before <- values(elev_clip)
vals_eth <- vals_before[!is.na(vals_before)]
#  Z-SCORE STANDARDIZATION
mean_val <- mean(vals_eth)
sd_val <- sd(vals_eth)
vals_std_eth <- (vals_eth - mean_val) / sd_val
vals_std_full <- rep(NA, length(vals_before))
vals_std_full[!is.na(vals_before)] <- vals_std_eth
values(elev_clip) <- vals_std_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1)\n")
#  GEOMETRY CHECK
geom_check <- compareGeom(elev_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")
#  SAVE FINAL RASTER (no year in filename)
output_filename <- "ethiopia_5km_elevation_zscore.tif"
output_path <- file.path("D:/adjusted_covaraite/constant", output_filename)
# writeRaster(elev_clip, filename=output_path, overwrite=TRUE, gdal="COMPRESS=LZW")
# cat("Raster saved:", output_path, "\n")
#  METADATA
metadata <- list(
  variable_name = "Elevation",
  variable_code = "elev",
  temporal_constancy = "constant across years (2000-2023)",
  country = "Ethiopia",
  source = "SRTM / GMTED2010",  # adjust as needed
  original_units = "meters",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(elev_clip)),
  resolution_m = paste(res(elev_clip), collapse=" x "),
  n_cells = ncell(elev_clip),
  n_cells_valid = length(vals_eth),
  percent_zeros_original = 0,                      # from diagnosis table
  skewness_original = 0.633105986,
  transformation = "none",
  standardization = "z-score",
  final_min = min(vals_eth),
  final_max = max(vals_eth),
  final_mean = mean(vals_eth),
  final_sd = sd(vals_eth),
  final_skewness = e1071::skewness(vals_eth),
  final_kurtosis = e1071::kurtosis(vals_eth),
  alignment_passed = geom_check,
  missing_final = sum(is.na(vals_before))
)

# Save metadata in same constant folder
output_path1 <- file.path("D:/metadara/constant", output_filename)
if(!dir.exists("D:/metadara/constant")) dir.create("D:/metadara/constant", recursive=TRUE)
jsonlite::write_json(metadata, path=gsub(".tif$","_metadata.json",output_path1), 
                     pretty=TRUE, auto_unbox=TRUE)
```

##  Lower Respiratory Infection

```{r}
#  LOAD RAW LRI RASTER (adjust year as needed, e.g., 2000)
lri_raw <- rast("D:/U5M/Covaraite original/LRI/LRI_ETH_2000.tif")  # adjust filename
if(nlyr(lri_raw) > 1) lri_raw <- lri_raw[[1]]
#  PROJECT + RESAMPLE TO 5KM GRID (use NEAREST to preserve zeros)
lri_proj <- if(!same.crs(lri_raw, template)) project(lri_raw, template) else lri_raw
lri_5km <- resample(lri_proj, template, method="near")
#  CROP + MASK TO ETHIOPIA
lri_crop <- terra::crop(lri_5km, ethiopia_poly_utm)
lri_clip <- terra::mask(lri_crop, ethiopia_poly_utm)
vals_before <- values(lri_clip)
vals_eth <- vals_before[!is.na(vals_before)]
#  Z-SCORE STANDARDIZATION
mean_val <- mean(vals_eth)
sd_val <- sd(vals_eth)
vals_std_eth <- (vals_eth - mean_val) / sd_val
vals_std_full <- rep(NA, length(vals_before))
vals_std_full[!is.na(vals_before)] <- vals_std_eth
values(lri_clip) <- vals_std_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1)\n")
#  GEOMETRY CHECK
geom_check <- compareGeom(lri_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")
#  SAVE FINAL RASTER
output_filename <- "ethiopia_5km_lri_2000_zscore.tif"
output_path <- file.path("D:/adjusted_covaraite/2000", output_filename)
# writeRaster(lri_clip, filename=output_path, overwrite=TRUE, gdal="COMPRESS=LZW")
# cat("Raster saved:", output_path, "\n")
#  METADATA
metadata <- list(
  variable_name = "Lower Respiratory Infection (LRI) prevalence",
  variable_code = "LRI",
  year = 2000,                                   # adjust per file
  country = "Ethiopia",
  source = "IHME / Global Burden of Disease",
  original_units = "cases per 1000? (adjust as needed)",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(lri_clip)),
  resolution_m = paste(res(lri_clip), collapse=" x "),
  n_cells = ncell(lri_clip),
  n_cells_valid = length(vals_eth),
  percent_zeros_original = 416495941,          # from diagnosis table
  skewness_original = 0.323522626,
  transformation = "none",
  standardization = "z-score",
  final_min = min(vals_eth),
  final_max = max(vals_eth),
  final_mean = mean(vals_eth),
  final_sd = sd(vals_eth),
  final_skewness = e1071::skewness(vals_eth),
  final_kurtosis = e1071::kurtosis(vals_eth),
  alignment_passed = geom_check,
  missing_final = sum(is.na(vals_before))
)
output_path1 <- file.path("D:/metadara/2000", output_filename)
```

##  Stunting

```{r}
#  LOAD RAW STUNTING RASTER (adjust year as needed, e.g., 2000)
stunting_raw <- rast("D:/stunt 2000.tif")  # adjust filename
if(nlyr(stunting_raw) > 1) stunting_raw <- stunting_raw[[1]]
#  PROJECT + RESAMPLE TO 5KM GRID (use NEAREST to preserve zeros)
stunting_proj <- if(!same.crs(stunting_raw, template)) project(stunting_raw, template) else stunting_raw
stunting_5km <- resample(stunting_proj, template, method="near")
#  CROP + MASK TO ETHIOPIA
stunting_crop <- terra::crop(stunting_5km, ethiopia_poly_utm)
stunting_clip <- terra::mask(stunting_crop, ethiopia_poly_utm)
vals_before <- values(stunting_clip)
vals_eth <- vals_before[!is.na(vals_before)]
#  Z-SCORE STANDARDIZATION
mean_val <- mean(vals_eth)
sd_val <- sd(vals_eth)
vals_std_eth <- (vals_eth - mean_val) / sd_val
vals_std_full <- rep(NA, length(vals_before))
vals_std_full[!is.na(vals_before)] <- vals_std_eth
values(stunting_clip) <- vals_std_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1)\n")
#  GEOMETRY CHECK
geom_check <- compareGeom(stunting_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")
#  SAVE FINAL RASTER
output_filename <- "ethiopia_stunting_2000_5km_zscore.tif"
output_path <- file.path("D:/adjusted_covaraite", output_filename)
# writeRaster(stunting_clip, filename=output_path, overwrite=TRUE, gdal="COMPRESS=LZW")
# cat("Raster saved:", output_path, "\n")
#  METADATA
metadata <- list(
  variable_name = "Stunting prevalence in children under 5",
  variable_code = "Stunting",
  year = 2000,                                   # adjust per file
  country = "Ethiopia",
  source = "DHS / UNICEF",
  original_units = "percent",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(stunting_clip)),
  resolution_m = paste(res(stunting_clip), collapse=" x "),
  n_cells = ncell(stunting_clip),
  n_cells_valid = length(vals_eth),
  percent_zeros_original = 414349097,          # from diagnosis table
  skewness_original = 0.162432268,
  transformation = "none",
  standardization = "z-score",
  final_min = min(vals_eth),
  final_max = max(vals_eth),
  final_mean = mean(vals_eth),
  final_sd = sd(vals_eth),
  final_skewness = e1071::skewness(vals_eth),
  final_kurtosis = e1071::kurtosis(vals_eth),
  alignment_passed = geom_check,
  missing_final = sum(is.na(vals_before))
)
output_path1 <- file.path("D:/metadara/2000", output_filename)
```

## Temperature (2m) 

```{r}
#  LOAD RAW TEMPERATURE RASTER (adjust year as needed)
temp_raw <- rast("D:/U5M/Covaraite original/temp_2m/temp_2000.tif")  # adjust filename
if(nlyr(temp_raw) > 1) temp_raw <- temp_raw[[1]]
#  PROJECT + RESAMPLE TO 5KM GRID (bilinear for continuous data)
temp_proj <- if(!same.crs(temp_raw, template)) project(temp_raw, template) else temp_raw
temp_5km <- resample(temp_proj, template, method="bilinear")
#  CROP + MASK TO ETHIOPIA
temp_crop <- terra::crop(temp_5km, ethiopia_poly_utm)
temp_clip <- terra::mask(temp_crop, ethiopia_poly_utm)
vals_before <- values(temp_clip)
vals_eth <- vals_before[!is.na(vals_before)]
#  Z-SCORE STANDARDIZATION
mean_val <- mean(vals_eth)
sd_val <- sd(vals_eth)
vals_std_eth <- (vals_eth - mean_val) / sd_val
vals_std_full <- rep(NA, length(vals_before))
vals_std_full[!is.na(vals_before)] <- vals_std_eth
values(temp_clip) <- vals_std_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1)\n")
#  GEOMETRY CHECK
geom_check <- compareGeom(temp_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")
#  SAVE FINAL RASTER
output_filename <- "ethiopia_5km_temp2m_2000_zscore.tif"
output_path <- file.path("D:/adjusted_covaraite/2000", output_filename)
#  METADATA
metadata <- list(
  variable_name = "2m Temperature",
  variable_code = "temp2m",
  year = 2000,                                   # adjust per file
  country = "Ethiopia",
  source = "ERA5",
  source_doi = "24381/cds.adbb2d47",
  original_units = "Kelvin",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(temp_clip)),
  resolution_m = paste(res(temp_clip), collapse=" x "),
  n_cells = ncell(temp_clip),
  n_cells_valid = length(vals_eth),
  percent_zeros_original = 0,                    # from diagnosis table
  skewness_original = -0.350254785,
  transformation = "none",
  standardization = "z-score",
  final_min = min(vals_eth),
  final_max = max(vals_eth),
  final_mean = mean(vals_eth),
  final_sd = sd(vals_eth),
  final_skewness = e1071::skewness(vals_eth),
  final_kurtosis = e1071::kurtosis(vals_eth),
  alignment_passed = geom_check,
  missing_final = sum(is.na(vals_before))
)
output_path1 <- file.path("D:/metadara/2000", output_filename)
```

##  Total Precipitation 

```{r}
#  LOAD RAW TOTAL PRECIPITATION
precip_raw <- rast("D:/presi_2000.tif")
if(nlyr(precip_raw) > 1) precip_raw <- precip_raw[[1]]
#  PROJECT + RESAMPLE (bilinear)
precip_proj <- if(!same.crs(precip_raw, template)) project(precip_raw, template) else precip_raw
precip_5km <- resample(precip_proj, template, method="bilinear")
#  CROP + MASK
precip_crop <- terra::crop(precip_5km, ethiopia_poly_utm)
precip_clip <- terra::mask(precip_crop, ethiopia_poly_utm)
vals_before <- values(precip_clip)
vals_eth <- vals_before[!is.na(vals_before)]
cat("Clipped to Ethiopia, valid pixels:", length(vals_eth), "\n")
#  LOG1P + Z-SCORE + CLIP TO [-5,5]
vals_log <- log1p(vals_eth)
mean_val <- mean(vals_log)
sd_val <- sd(vals_log)
vals_std_eth <- (vals_log - mean_val) / sd_val
vals_std_eth <- pmax(pmin(vals_std_eth, 5), -5)   # cap at ±5
vals_std_full <- rep(NA, length(vals_before))
vals_std_full[!is.na(vals_before)] <- vals_std_eth
values(precip_clip) <- vals_std_full
cat("Log1p + z-score + clipping to [-5,5] complete\n")
#  GEOMETRY CHECK
geom_check <- compareGeom(precip_clip, template, stopOnError=FALSE)
cat("Geometry alignment:", ifelse(geom_check,"PASS","FAIL"), "\n")
#  SAVE
output_filename <- "ethiopia_5km_totalprecip_2000_log1p_zscore.tif"
output_path <- file.path("D:/adjusted_covaraite", output_filename)
# cat("Raster saved:", output_path, "\n")
#  METADATA (add clipping info)
metadata <- list(
  variable_name = "Total Precipitation",
  variable_code = "total_precip",
  year = 2000,
  country = "Ethiopia",
  source = "ERA5",
  source_doi = "24381/cds.adbb2d47",
  original_units = "m",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(precip_clip)),
  resolution_m = paste(res(precip_clip), collapse=" x "),
  n_cells = ncell(precip_clip),
  n_cells_valid = length(vals_eth),
  percent_zeros_original = 0,
  skewness_original = 563184314,
  transformation = "log1p",
  standardization = "z-score",
  clipping = "[-5, 5]",
  final_min = min(vals_std_eth),
  final_max = max(vals_std_eth),
  final_mean = mean(vals_std_eth),
  final_sd = sd(vals_std_eth),
  final_skewness = e1071::skewness(vals_std_eth),
  final_kurtosis = e1071::kurtosis(vals_std_eth),
  transformed_skewness = e1071::skewness(vals_log),
  transformed_kurtosis = e1071::kurtosis(vals_log),
  alignment_passed = geom_check,
  missing_final = sum(is.na(vals_before))
)
output_path1 <- file.path("D:/metadara/2000", output_filename)
```

##  Volumetric Soil Water

```{r}
#  LOAD RAW VOLUMETRIC SOIL WATER RASTER (adjust year as needed)
#    Note: This is for the first soil layer (0-7 cm). Adjust filename accordingly.
soilwater_raw <- rast("D:/vol2_2000.tif")
if(nlyr(soilwater_raw) > 1) soilwater_raw <- soilwater_raw[[1]]
#  PROJECT + RESAMPLE TO 5KM GRID (bilinear for continuous data)
soilwater_proj <- if(!same.crs(soilwater_raw, template)) project(soilwater_raw, template) else soilwater_raw
soilwater_5km <- resample(soilwater_proj, template, method="bilinear")
#  CROP + MASK TO ETHIOPIA
soilwater_crop <- terra::crop(soilwater_5km, ethiopia_poly_utm)
soilwater_clip <- terra::mask(soilwater_crop, ethiopia_poly_utm)
vals_before <- values(soilwater_clip)
vals_eth <- vals_before[!is.na(vals_before)]
cat("Clipped to Ethiopia, valid pixels:", length(vals_eth), "\n")
cat("  (Note: No zeros, no transform applied, following diagnosis)\n")
#  Z-SCORE STANDARDIZATION (no transformation)
mean_val <- mean(vals_eth)
sd_val <- sd(vals_eth)
vals_std_eth <- (vals_eth - mean_val) / sd_val
# Optional: clip to [-5,5] if extreme outliers exist (but skewness is very low)
# vals_std_eth <- pmax(pmin(vals_std_eth, 5), -5)
vals_std_full <- rep(NA, length(vals_before))
vals_std_full[!is.na(vals_before)] <- vals_std_eth
values(soilwater_clip) <- vals_std_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1)\n")
#  GEOMETRY CHECK
geom_check <- compareGeom(soilwater_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")
#  SAVE FINAL RASTER
output_filename <- "ethiopia_5km_soilwater_2000_zscore.tif"
output_path <- file.path("D:/adjusted_covaraite/2000", output_filename)
#  METADATA
metadata <- list(
  variable_name = "Volumetric Soil Water (Layer 1, 0-7 cm)",
  variable_code = "soilwater",
  year = 2000,                                   # adjust per file
  country = "Ethiopia",
  source = "ERA5-Land",
  source_doi = "24381/cds.68d2bb30",
  original_units = "m³/m³",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(soilwater_clip)),
  resolution_m = paste(res(soilwater_clip), collapse=" x "),
  n_cells = ncell(soilwater_clip),
  n_cells_valid = length(vals_eth),
  percent_zeros_original = 0,                    # from diagnosis table
  skewness_original = 0.01681263,
  transformation = "none",
  standardization = "z-score",
  final_min = min(vals_eth),
  final_max = max(vals_eth),
  final_mean = mean(vals_eth),
  final_sd = sd(vals_eth),
  final_skewness = e1071::skewness(vals_eth),
  final_kurtosis = e1071::kurtosis(vals_eth),
  alignment_passed = geom_check,
  missing_final = sum(is.na(vals_before))
)
output_path1 <- file.path("D:/metadara/2000", output_filename)
```

##  Volumetric Soil Water 1 

```{r}
#  LOAD RAW VOLUMETRIC SOIL WATER 1 RASTER (adjust year and path)
#    This is typically the second soil layer (e.g., 7-28 cm). Adjust filename.
soilwater1_raw <- rast("D:/vol1_2000.tif")
if(nlyr(soilwater1_raw) > 1) soilwater1_raw <- soilwater1_raw[[1]]
cat("Raw Volumetric Soil Water 1 raster loaded\n")
#  PROJECT + RESAMPLE TO 5KM GRID (bilinear for continuous data)
soilwater1_proj <- if(!same.crs(soilwater1_raw, template)) project(soilwater1_raw, template) else soilwater1_raw
soilwater1_5km <- resample(soilwater1_proj, template, method="bilinear")
#  CROP + MASK TO ETHIOPIA
soilwater1_crop <- terra::crop(soilwater1_5km, ethiopia_poly_utm)
soilwater1_clip <- terra::mask(soilwater1_crop, ethiopia_poly_utm)
vals_before <- values(soilwater1_clip)
vals_eth <- vals_before[!is.na(vals_before)]
cat("Clipped to Ethiopia, valid pixels:", length(vals_eth), "\n")
cat("  (Note: No zeros, no transform applied – skewness = 0.073, near-symmetric)\n")
#  Z-SCORE STANDARDIZATION (no transformation)
mean_val <- mean(vals_eth)
sd_val <- sd(vals_eth)
vals_std_eth <- (vals_eth - mean_val) / sd_val
# Optional clipping (if extreme outliers exist, uncomment next line)
# vals_std_eth <- pmax(pmin(vals_std_eth, 5), -5)
vals_std_full <- rep(NA, length(vals_before))
vals_std_full[!is.na(vals_before)] <- vals_std_eth
values(soilwater1_clip) <- vals_std_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1)\n")
#  GEOMETRY CHECK
geom_check <- compareGeom(soilwater1_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")
#  SAVE FINAL RASTER
output_filename <- "ethiopia_5km_soilwater1_2000_zscore.tif"
output_path <- file.path("D:/adjusted_covaraite", output_filename)
# writeRaster(soilwater1_clip, filename=output_path, overwrite=TRUE, gdal="COMPRESS=LZW")
# cat("Raster saved:", output_path, "\n")

#  METADATA
metadata <- list(
  variable_name = "Volumetric Soil Water (Layer 2, deeper)",
  variable_code = "soilwater1",
  year = 2000,
  country = "Ethiopia",
  source = "ERA5-Land",
  source_doi = "24381/cds.68d2bb30",
  original_units = "m³/m³",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(soilwater1_clip)),
  resolution_m = paste(res(soilwater1_clip), collapse=" x "),
  n_cells = ncell(soilwater1_clip),
  n_cells_valid = length(vals_eth),
  percent_zeros_original = 0,
  skewness_original = 0.073210556,
  transformation = "none",
  standardization = "z-score",
  final_min = min(vals_eth),
  final_max = max(vals_eth),
  final_mean = mean(vals_eth),
  final_sd = sd(vals_eth),
  final_skewness = e1071::skewness(vals_eth),
  final_kurtosis = e1071::kurtosis(vals_eth),
  alignment_passed = geom_check,
  missing_final = sum(is.na(vals_before))
)
output_path1 <- file.path("D:/metadara/2000", output_filename)
```

## creat refference for marge

```{r}
# Create a raster of cell numbers
cell_raster <- temp
values(cell_raster) <- 1:ncell(cell_raster)
cell_raster[is.na(temp)] <- NA   # mask out outside Ethiopia
# Convert to data frame (only valid cells)
ref_df <- as.data.frame(cell_raster, xy = TRUE, na.rm = FALSE)
names(ref_df) <- c("x", "y", "cell_id")
# Add consecutive pixel ID and coordinates
ref_df$pixel_id <- 1:nrow(ref_df)
utm_crs <- crs(temp)
ll_crs <- "EPSG:4326"
coords_utm <- cbind(ref_df$x, ref_df$y)
coords_ll <- project(coords_utm, from = utm_crs, to = ll_crs)
ref_df$longitude <- coords_ll[,1]
ref_df$latitude <- coords_ll[,2]
# Save reference
write.csv(ref_df, "D:/pixel_reference_5km.csv", row.names = FALSE)
```

## praper for marge

## Covariate preprocessing

Twenty‑five raster covariates (2000‑2015) were standardised to consistent names and merged into a single data frame. For each year, pixel values were extracted using a reference grid, column names were renamed (e.g., `stunting_2000...` → `Stunting`), a `year` column was added, and the result was saved. All yearly files were then row‑bound. The final dataset contains 45,931 pixels × 16 years × 25 covariates. Code: `R/01_prepare_covariates.R`.

```{r}
library(terra)
ref_df <- read.csv("D:/pixel_reference_5km.csv")
folder <- "D:/adjusted_covaraite"
tif_files <- list.files(folder, pattern = "\\.tif$", full.names = TRUE)
for (f in tif_files) {
  r <- rast(f)
  valid <- global(r, "notNA", na.rm = FALSE)[1,1]
  cat(basename(f), ": valid pixels =", valid, "\n")
}
# Load reference table (must have columns: cell_id, pixel_id, longitude, latitude)
ref_df <- read.csv("D:/pixel_reference_5km.csv")
folder_2000 <- "D:/adjusted_covaraite"
tif_files <- list.files(folder_2000, pattern = "\\.tif$", full.names = TRUE)
clean_name <- function(f) {
  gsub("_[0-9]{4}_5km_zscore_capped\\.tif$", "", basename(f))
}
result <- ref_df[, c("pixel_id", "longitude", "latitude")]
for (f in tif_files) {
  r <- rast(f)
  vname <- clean_name(f)
  result[[vname]] <- r[ref_df$cell_id]
}
result$year <- 2000
# After merging, check valid counts per covariate
cov_cols <- setdiff(names(result), c("pixel_id", "longitude", "latitude", "year"))
valid_counts <- sapply(result[, cov_cols], function(x) sum(!is.na(x)))
print(valid_counts)
# Define mapping for 2000 files (replace "2001" with "2000")
name_mapping_2000 <- c(
  "ethiopia_deareaha_2000_5km_zscore.tif" = "deareha",
  "ethiopia_dewpoint_temp_2m_2000_5km_zscore.tif" = "dewpoint_temp_2m",
  "ethiopia_dpt3_2000_5km_zscore.tif" = "DPT3",
  "ethiopia_education_2000_5km_zscore.tif" = "Education",
  "ethiopia_5km_chirps_2000_zscore.tif" = "chrips",
  "ethiopia_5km_ebf_2000_zscore.tif" = "EBF",
  "ethiopia_5km_elevation_zscore.tif" = "Elevation",
  "ethiopia_5km_lri_2000_zscore.tif" = "LRI",
  "ethiopia_slope_5km_zscore.tif" = "Slope",
  "ethiopia_5km_soilwater_2000_zscore.tif" = "Volumetric_Soil_Water",
  "ethiopia_5km_soilwater1_2000_zscore.tif" = "Volumetric_Soil_Water_1",
  "ethiopia_stunting_2000_5km_zscore.tif" = "Stunting",
  "ethiopia_5km_temp2m_2000_zscore.tif" = "Temperature_2m",
  "ethiopia_5km_totalprecip_2000_log1p_zscore.tif" = "Total_Precipitation",
  "ethiopia_hiv_2000_5km_zscore.tif" = "HIV",
  "ethiopia_malaria_2000_5km_zlog.tif" = "malaria",
  "ethiopia_ndvi_2000_5km_zscore.tif" = "NDVI",
  "ethiopia_nighttime_light_2000_5km_binary.tif" = "Night_time_light",
  "ethiopia_pm2_5_2000_5km_zscore.tif" = "PM5",
  "ethiopia_potential_evaporation_2000_5km_zscore.tif" = "potentialevaporation",
  "ethiopia_runoff_2000_5km_zscore.tif" = "runoff",
  "ethiopia_severe_wasting_2000_5km_zlog.tif" = "Severe_Wasting",
  "ethiopia_solar_radiation_2000_5km_zscore.tif" = "solarradiationdown",
  "ethiopia_wash_2000_5km_zlog.tif" = "WASH",
  "ethiopia_water_2000_5km_zscore.tif" = "Water"
)
# Rename columns in 'result' data frame (2000)
names(result) <- ifelse(names(result) %in% names(name_mapping_2000), 
                        name_mapping_2000[names(result)], 
                        names(result))
# Check new column names
names(result)
saveRDS(result, "D:/covariates_2000.rds")
```


```{r}
y00<-readRDS("D:/covariates_2000.rds")
y01<-readRDS("D:/covariates_2001.rds")
y02<-readRDS("D:/covariates_2002.rds")
y03<-readRDS("D:/covariates_2003.rds")
y04<-readRDS("D:/covariates_2004.rds")
y05<-readRDS("D:/covariates_2005.rds")
y06<-readRDS("D:/covariates_2006.rds")
y07<-readRDS("D:/covariates_2007.rds")
y08<-readRDS("D:/covariates_2008.rds")
y09<-readRDS("D:/covariates_2009.rds")
y10<-readRDS("D:/covariates_2010.rds")
y11<-readRDS("D:/covariates_2011.rds")
y12<-readRDS("D:/covariates_2012.rds")
y13<-readRDS("D:/covariates_2013.rds")
y14<-readRDS("D:/covariates_2014.rds")
y15<-readRDS("D:/covariates_2015.rds")
names(y00)
names(y01)
names(y15)
```

```{r}
all_years <- rbind(y00, y01, y02, y03, y04, y05, y06, y07, y08,
                   y09, y10, y11, y12, y13, y14, y15)
```

```{r}
names(all_years)
```

```{r}
saveRDS(all_years, "D:/all_years.rds")
```

```{r}
all_years<-readRDS("D:/all_years.rds")
# Convert all_years to a plain data frame (if not already)
all_years <- as.data.frame(all_years)

# Convert all covariate columns to numeric (in case they are lists or factors)
cov_names <- c("deareha", "dewpoint_temp_2m", "DPT3", "Education", "chrips",  "EBF", "Elevation", "LRI", "Slope", "Volumetric_Soil_Water",
    "Volumetric_Soil_Water_1", "Stunting", "Temperature_2m",
               "Total_Precipitation", "HIV", "malaria", "NDVI", "Night_time_light",  "PM5", "potentialevaporation", "runoff", "Severe_Wasting", "solarradiationdown", "WASH", "Water")
for (col in cov_names) {
  if (is.list(all_years[[col]])) {
    all_years[[col]] <- unlist(all_years[[col]])
  }
  all_years[[col]] <- as.numeric(all_years[[col]])
}
# Now subset to complete cases
df_complete <- all_years[complete.cases(all_years[, cov_names]), ]
# Check that df_complete has rows
nrow(df_complete)
```

```{r}
cov_names <- c(
  "deareha", "dewpoint_temp_2m", "DPT3", "Education", "chrips",
  "EBF", "Elevation", "LRI", "Slope", "Volumetric_Soil_Water",
  "Volumetric_Soil_Water_1", "Stunting", "Temperature_2m",
  "Total_Precipitation", "HIV", "malaria", "NDVI", "Night_time_light",
  "PM5", "potentialevaporation", "runoff", "Severe_Wasting",
  "solarradiationdown", "WASH", "Water"
)
#  Subset to complete cases (rows without any NA in these covariates)
df_complete <- all_years[complete.cases(all_years[, cov_names]), ]
#  Compute full correlation matrix (Pearson)
cor_matrix_full <- cor(df_complete[, cov_names], method = "pearson")
# Save full correlation matrix as CSV (optional)
write.csv(cor_matrix_full, "D:/full_correlation_matrix_all.csv", row.names = TRUE)
#  Identify variables to drop where |r| > 0.7
to_drop <- findCorrelation(cor_matrix_full, cutoff = 0.7, exact = TRUE, names = TRUE)
cat("Variables to drop (|r| > 0.7):\n", to_drop, "\n")
#  Keep only the uncorrelated covariates
selected_covs <- setdiff(cov_names, to_drop)
cat("\nFinal selected covariates (|r| ≤ 0.7):\n", selected_covs, "\n")
# Save the list as a text file
writeLines(selected_covs, "D:/selected_covariates_list.txt")
#  Correlation matrix of the selected covariates only
cor_selected <- cor(df_complete[, selected_covs], method = "pearson")
# Save reduced correlation matrix to CSV
write.csv(cor_selected, "D:/reduced_correlation_selected.csv", row.names = TRUE)
#  Plot the correlation matrix of the selected covariates
#  Show on screen
windows()
corrplot(cor_selected, method = "color", type = "upper", order = "hclust",
         tl.cex = 0.9, tl.srt = 45, tl.col = "black",
         addCoef.col = "black", number.cex = 0.65, diag = FALSE,
         title = "Selected Covariates")

#  Save to TIFF (no window will pop up during this step)
tiff("D:/correlation_plot_selected.tiff", width = 14, height = 12, units = "in", res = 300)
corrplot(cor_selected, method = "color", type = "upper", order = "hclust",
         tl.cex = 0.9, tl.srt = 45, tl.col = "black",
         addCoef.col = "black", number.cex = 0.65, diag = FALSE,
         title = "Selected Covariates", mar = c(0,0,2,0))
dev.off()
```

# Under five population

```{r}
eth_shp <- vect("D:/gadm41_ETH_0.shp") 
target_crs <- "EPSG:32637"
eth_utm <- project(eth_shp, target_crs)
e <- ext(eth_utm)
template_5km <- rast(e, resolution = 5000, crs = target_crs)
values(template_5km) <- 1:ncell(template_5km)
template_5km <- mask(template_5km, eth_utm)
template <- mask(template_5km, eth_utm)
```

```{r worldpop}
# WORLD POPULATION (UNDER-5) AGGREGATION TO 5km TEMPLATE
# For under-five mortality rate calculation (denominator)
#  Load libraries and template
library(terra)
library(sf)
library(exactextractr)
library(jsonlite)
library(e1071)
pop_100m_path <- "D:/Pop_2000.tif"
output_dir <- "D:/Adjusted_Raster"
output_file <- file.path(output_dir, "Pop2000_5km.tif")
metadata_file <- file.path(output_dir, "Pop2000_5km_metadata.json")
diagnostic_file <- file.path(output_dir, "Pop2000_5km_diagnostics.csv")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
#  Load template and 100m population
ref <- rast(template)
pop_100m <- rast(pop_100m_path)
#  Convert template to polygons (in same CRS as population)
ref_wgs <- project(ref, crs(pop_100m))
polys_wgs <- as.polygons(ref_wgs, dissolve = FALSE)
ref_sf <- st_as_sf(polys_wgs)
#  Aggregate using exact_extract (sum of population per 5km cell)
pop_sum_values <- exact_extract(pop_100m, ref_sf, fun = "sum")
#  Create 5km population raster
pop_5km <- ref
values(pop_5km)[!is.na(values(pop_5km))] <- pop_sum_values
#  Diagnostics
pop_vals <- values(pop_5km)
pop_vals_nonNA <- pop_vals[!is.na(pop_vals)]
diagnostics <- data.frame(
  variable = "Pop2000_5km",
  year = 2000,
  source = "WorldPop",
  original_resolution = "100m",
  aggregation_method = "sum (exact_extract)",
  n_cells_total = ncell(pop_5km),
  n_cells_valid = length(pop_vals_nonNA),
  n_cells_missing = sum(is.na(pop_vals)),
  min = min(pop_vals_nonNA, na.rm = TRUE),
  max = max(pop_vals_nonNA, na.rm = TRUE),
  sum = sum(pop_vals_nonNA, na.rm = TRUE),
  mean = mean(pop_vals_nonNA, na.rm = TRUE),
  sd = sd(pop_vals_nonNA, na.rm = TRUE),
  skewness = skewness(pop_vals_nonNA, na.rm = TRUE),
  percent_zeros = 100 * sum(pop_vals_nonNA == 0) / length(pop_vals_nonNA),
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"]
)
write.csv(diagnostics, diagnostic_file, row.names = FALSE)
cat("Diagnostics saved to:", diagnostic_file, "\n")
#  Metadata JSON
metadata <- list(
  variable_name = "Under-five population count",
  variable_code = "Pop2000_5km",
  year = 2000,
  country = "Ethiopia",
  source = "WorldPop",
  source_url = "https://www.worldpop.org/",
  original_units = "persons per 100m grid cell",
  final_units = "persons per 5km grid cell",
  aggregation = "sum of 100m cells within each 5km cell",
  spatial_resolution_m = paste(res(pop_5km), collapse = " x "),
  crs = as.character(crs(pop_5km)),
  n_cells_total = ncell(pop_5km),
  n_cells_valid = length(pop_vals_nonNA),
  n_cells_missing = sum(is.na(pop_vals)),
  total_population = sum(pop_vals_nonNA, na.rm = TRUE),
  mean_cell_population = mean(pop_vals_nonNA, na.rm = TRUE),
  median_cell_population = median(pop_vals_nonNA, na.rm = TRUE),
  min_cell_population = min(pop_vals_nonNA, na.rm = TRUE),
  max_cell_population = max(pop_vals_nonNA, na.rm = TRUE),
  skewness = skewness(pop_vals_nonNA, na.rm = TRUE),
  processing_date = as.character(Sys.Date()),
  software = list(R_version = R.version.string, terra_version = as.character(packageVersion("terra")))
)
write_json(metadata, metadata_file, pretty = TRUE, auto_unbox = TRUE)
#  Save raster
writeRaster(pop_5km, output_file, gdal = "COMPRESS=LZW", overwrite = TRUE)
```

```{r inhabited areas}
# INHABITED AREAS MASK (5km) – ETHIOPIA 2000
# Based on: NDVI >= 0.1, land (not water), population >= 1
library(terra)
library(sf)
library(jsonlite)
library(e1071)
ndvi_path <- "D:/U5M/NDVI Ethiopia/NDVI_2000_Ethiopia.tif"
water_path <- "D:/shap/ethio waterbody.tif"
pop_path <- "D:/U5M/population/Pop00_5km.tif"   # use the newly created 5km population
output_dir <- "D:/Adjusted_Raster"
output_mask <- file.path(output_dir, "inhabit00_5km.tif")
metadata_file <- file.path(output_dir, "inhabit00_5km_metadata.json")
diagnostic_file <- file.path(output_dir, "inhabit00_5km_diagnostics.csv")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
#  Load template
template <- rast(template)
#  NDVI mask (>= 0.1)
ndvi_raw <- rast(ndvi_path)
if (nlyr(ndvi_raw) > 1) ndvi_raw <- ndvi_raw[[1]]
# Scale if MODIS (0-10000 scale → 0-1)
if (minmax(ndvi_raw)[2] > 1) ndvi_raw <- ndvi_raw * 0.0001
ndvi_proj <- project(ndvi_raw, template, method = "bilinear")
ndvi_mask <- ifel(ndvi_proj >= 0.1, 1, NA)
cat("NDVI mask: inhabited cells =", sum(values(ndvi_mask) == 1, na.rm = TRUE), "\n")
#  Water mask (0 = land, 1 = water)
cat("\n--- Water mask ---\n")
water <- rast(water_path)[[1]]
water_proj <- project(water, template, method = "near")
land_mask <- ifel(water_proj == 0, 1, NA)   # 1 = land, NA = water
cat("Land mask: land cells =", sum(values(land_mask) == 1, na.rm = TRUE), "\n")
#  Population mask (>= 1 person)
cat("\n--- Population mask ---\n")
pop_5km <- rast(pop_path)
pop_mask <- ifel(pop_5km >= 1, 1, NA)
cat("Population mask: inhabited cells =", sum(values(pop_mask) == 1, na.rm = TRUE), "\n")

#  Align all masks (resample to template to be safe)
ndvi_mask <- resample(ndvi_mask, template, method = "near")
land_mask <- resample(land_mask, template, method = "near")
pop_mask  <- resample(pop_mask,  template, method = "near")

#  Combine (logical AND)
inhabit_mask <- app(c(ndvi_mask, land_mask, pop_mask), fun = function(x) {
  if (all(x == 1, na.rm = TRUE)) 1 else NA
})
cat("Final inhabited mask: cells =", sum(values(inhabit_mask) == 1, na.rm = TRUE), "\n")
#  Diagnostics
mask_vals <- values(inhabit_mask)
total_cells <- sum(!is.na(values(template)))
inhabited_cells <- sum(mask_vals == 1, na.rm = TRUE)
diagnostics <- data.frame(
  variable = "inhabit00_5km",
  year = 2000,
  method = "NDVI >= 0.1 AND land (water == 0) AND population >= 1",
  total_5km_cells = total_cells,
  inhabited_cells = inhabited_cells,
  percent_inhabited = 100 * inhabited_cells / total_cells,
  ndvi_threshold = 0.1,
  water_mask = "land only (water == 0)",
  population_threshold = ">= 1 person",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"]
)
write.csv(diagnostics, diagnostic_file, row.names = FALSE)
cat("Diagnostics saved to:", diagnostic_file, "\n")
#  Metadata JSON
metadata <- list(
  variable_name = "Inhabited areas mask",
  variable_code = "inhabit00_5km",
  year = 2000,
  country = "Ethiopia",
  definition = "Cells classified as inhabited if: (1) NDVI >= 0.1, (2) land (not water), (3) population >= 1",
  data_sources = list(
    NDVI = "MODIS (scaled 0-1)",
    water = "Global Surface Water (ESA)",
    population = "WorldPop aggregated to 5km"
  ),
  spatial_resolution_m = paste(res(inhabit_mask), collapse = " x "),
  crs = as.character(crs(inhabit_mask)),
  n_cells_total = total_cells,
  n_cells_inhabited = inhabited_cells,
  percent_inhabited = round(100 * inhabited_cells / total_cells, 2),
  processing_steps = c(
    "Project all inputs to template CRS",
    "Resample to 5km template",
    "Create binary masks: NDVI>=0.1, land=1, pop>=1",
    "Combine with logical AND"
  ),
  processing_date = as.character(Sys.Date()),
  software = list(R_version = R.version.string, terra_version = as.character(packageVersion("terra")))
)
write_json(metadata, metadata_file, pretty = TRUE, auto_unbox = TRUE)
cat("Metadata saved to:", metadata_file, "\n")
#  Save final mask
writeRaster(inhabit_mask, output_mask, gdal = "COMPRESS=LZW", overwrite = TRUE)
```


